---
title: "Extending the pipeline"
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Extending the pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

In this vignette we will reuse the project initialized within the Quick start guide (`vignette("scdrake")`) that should
live in `~/scdrake_projects/pbmc1k`.

***

`{drake}`'s pipeline definitions (plans) are R objects which can be arbitrarily extended by new targets.
There are several ways how to extend plans provided by `{scdrake}`, but the easiest way is to utilize the built-in
plan extension called inside the `{drake}` init scripts `_drake_single_sample.R` and `_drake_integration.R`.

Before the plan (or more precisely, `drake::drake_config()`) is returned from those scripts, an additional R script
(`plan_custom.R` in project root by default) is sourced. In this script, the last returned value must be a valid
`drake` plan (`drake::drake_plan()`) which is consequently merged with the original plan.

The path to R script with custom plan is taken from the `SCDRAKE_PLAN_CUSTOM_FILE` environment variable.

***

# Defining a custom plan

Let's try a dummy additional plan which is already present in `plan_custom.R`. Just uncomment the lines under
`Example:`, or if you want, you can try to define your own target/s. A quick intro to `{drake}` plans can be found
[here](https://books.ropensci.org/drake/walkthrough.html) and
[here](https://books.ropensci.org/drake/plans.html#intro-to-plans).

We also need to tell `{drake}` to make this new target. Open `config/pipeline.yaml` and set `DRAKE_TARGETS` to
`["my_target"]` (or more targets if you have defined them).

Now just run the pipeline as usual:

```{r, eval = FALSE}
run_single_sample_r()
```

In the terminal you should see an informative text `ℹ Extending the plan with a custom one defined in 'plan_custom.R'`.

When the pipeline finishes, you can load the new target to your session with

```{r, eval = FALSE}
drake::loadd(my_target)
```

***

## Parametrized custom plan

In `plan_custom.R` you can use any variables defined in `_drake_single_sample.R` or `_drake_integration.R`.
Probably the most important are `cfg` and `cfg_pipeline` lists holding pipeline parameters.
Note that in `plan_custom.R` all variables from the parent script are locked and cannot be modified.

Let's see how you can utilize the `cfg` list. Open `config/single_sample/01_input_qc.yaml` and add a new line:

```yaml
MY_GENE: "NOC2L"
```

Then replace the code in `plan_custom.R` with

```{r, eval = FALSE}
drake::drake_plan(
  my_target = scater::plotExpression(sce_final_input_qc, cfg$input_qc$MY_GENE, exprs_values = "counts", swap_rownames = "SYMBOL")
)
```

You can see we used the `MY_GENE` parameter defined in the config file. Later, `{drake}` will replace
`cfg$input_qc$MY_GENE` with its value `"NOC2L"`.

***

In case you don't want to use the standard config files, you can make your own one, e.g. `config/my_params.yaml`:

```yaml
MY_GENE: "NOC2L"
```

, and use it in `plan_custom.R`:

```{r, eval = FALSE}
my_cfg <- load_config("config/my_params.yaml")
drake::drake_plan(
  my_target = scater::plotExpression(sce_final_input_qc, my_cfg$MY_GENE, exprs_values = "counts", swap_rownames = "SYMBOL")
)
```

***

# Extending the RMarkdown documents

## Stage-specific RMarkdown documents

All RMarkdown files used for stage reports are located in the `Rmd/` directory in project's root.
Feel free to modify them to your needs. Just keep in mind that when you call `update_project()`,
those files will be **overwritten** by the default ones. To overcome this situation, you can save your modified file
using a different name and then modify the parameter specifying a path to stage's Rmd file.

For example, you modify `Rmd/single_sample/01_input_qc.Rmd`, save it as `Rmd/single_sample/01_input_qc_modified.Rmd`,
and change accordingly the `INPUT_QC_REPORT_RMD_FILE` parameter in `config/single_sample/01_input_qc.yaml`.

## Custom RMarkdown documents

For additional RMarkdown documents you just need to incorporate a file-returning target into your custom plan, e.g.

`Rmd/my_report.Rmd`

````{verbatim, lang = "markdown"}
---
title: "My report"
---

```{r}
drake::loadd(sce_final_norm_clustering)
scater::plotReducedDim(sce_final_norm_clustering, dimred = "umap", colour_by = "cluster_graph_louvain")
```
````

`plan_custom.R`

```{r, eval = FALSE}
drake::drake_plan(
  my_report = drake::target(
    rmarkdown::render(
      drake::knitr_in("Rmd/my_report.Rmd"),
      output_file = here::here("my_report.html"),
      knit_root_dir = here::here()
    ),

    format = "file"
  )
)
```

Let's break down the plan above:

- `my_report` target has `format = "file"` meaning `{drake}` expects the return value to be a character vector
  of file or directory paths, and `rmarkdown::render()` returns path to output file.
- `drake::knitr_in("Rmd/my_report.Rmd")` is a special function which marks the Rmd file as a dependency.
  Internally, it scans active code chunks and search for calls to `drake::loadd()` and `drake::readd()`, and marks
  the targets inside as dependencies of the target (`my_report`).
- By default, `{knitr}`, which is responsible for rendering of Rmd files, uses working directory the same as the
  location of the Rmd file. This is violating our project-based approach (*everything is specified relative to project
  root*), and so we are using `here::here()` to specify the output file and working directory.
  `here::here()` remembers the project root directory on its load and converts root-relative paths to absolute.
  You can try it yourself: call `here::here()` or `here::here("Rmd/my_report.Rmd")`.

Just for curiosity, we can see the dependencies of `my_target` using
`drake::r_deps_target(my_report, source = "_drake_single_sample.R")`:

```
  name                      type       hash            
1 here::here                namespaced NA              
2 rmarkdown::render         namespaced NA              
3 sce_final_norm_clustering loadd      911414192d751378
4 Rmd/my_report.Rmd         knitr_in   NA
```

## Reusing the stage-specific RMarkdown documents

Another possibility is to reuse the machinery responsible for rendering of stage reports, that is:

- The headmost part of Rmd documents (you can check e.g. 
  [Rmd/single_sample/01_input_qc.Rmd](https://github.com/bioinfocz/scdrake/blob/main/inst/Rmd/single_sample/01_input_qc.Rmd)):

````{verbatim, lang = "markdown"}
---
title: "`r params$title`"
author: "Your name"
institute: "Your institute"
date: "`r glue::glue('Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z\")}')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: false
    theme: "flatly"
    self_contained: true
    code_download: true
    df_print: "paged"
params:
  css_file: !expr here::here("Rmd/common/stylesheet.css")
  drake_cache_dir: !expr here::here(".drake")
  title: "Your title"
css: "`r params$css_file`"
---

```{r, message = FALSE, warning = FALSE}
drake_cache_dir <- params_$drake_cache_dir
drake::loadd(your_target_1, your_target_2, ..., path = drake_cache_dir)
```
````

The example Rmd document above has several parameters:

- `css_file`: a path to CSS file with HTML styling.
- `drake_cache_dir`: used to load targets from nondefault cache directory.
- `title`: a dynamic document title.

***

The second part of the machinery is the `generate_stage_report()` function. Internally, it's a wrapper around
`rmarkdown::render()` with some sensible defaults, and passing `css_file`, `drake_cache_dir` and other user-defined
parameters to the Rmd document. For the Rmd document above, we would call this function as

```{r, eval = FALSE}
generate_stage_report(
  ## -- We assume that the document is saved here.
  "Rmd/my_report.Rmd",
  "output/my_report.html",
  params = list(title = "My report")
)
```

***

And here is an example of target rendering a report for the `01_input_qc` stage
([source](https://github.com/bioinfocz/scdrake/blob/main/R/plans_single_sample.R)):

```{r, eval = FALSE}
drake::drake_plan(
  report_input_qc = target(
      generate_stage_report(
        rmd_file = knitr_in(!!cfg$INPUT_QC_REPORT_RMD_FILE),
        out_html_file_name = file_out(!!cfg$INPUT_QC_REPORT_HTML_FILE),
        css_file = file_in(!!cfg_main$CSS_FILE),
        message = !!cfg$INPUT_QC_KNITR_MESSAGE,
        warning = !!cfg$INPUT_QC_KNITR_WARNING,
        echo = !!cfg$INPUT_QC_KNITR_ECHO,
        other_deps = list(
          file_in(!!here("Rmd/common/_header.Rmd")),
          file_in(!!here("Rmd/common/_footer.Rmd")),
          file_in(!!here("Rmd/single_sample/01_input_qc_children/empty_droplets.Rmd"))
        ),
        drake_cache_dir = !!cfg_pipeline$DRAKE_CACHE_DIR
      ),
      format = "file"
    )
)
```

You can see that almost all function parameters are dynamic, based on config file (`cfg`).
Also, we force `{drake}` to watch for changes in other child Rmd documents by specifying them inside `other_deps`.

Note that `drake::file_in()` is watching for changes in file size/structure and not for calls to `drake::loadd()` and
`drake::readd()` as `drake::knitr_in()` is doing (static code analysis).
