---
title: "FAQ & Howtos"
date: "`r Sys.Date()`"
author:
  - name: Jiri Novotny
    affiliation: Institute of Molecular Genetics of the Czech Academy of Sciences
    email: jiri.novotny@img.cas.cz
  - name: Jan Kubovciak
    affiliation: Institute of Molecular Genetics of the Czech Academy of Sciences
    email: jan.kubovciak@img.cas.cz
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{FAQ & Howtos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

# Are you planning to migrate the pipeline to `{targets}`?

We will [cite](https://books.ropensci.org/targets/drake.html) the author of both packages:

> `{targets}` is the successor of `{drake}`, an older pipeline tool. As of 2021-01-21, `{drake}` is superseded,
which means there are no plans for new features or discretionary enhancements, but basic maintenance and support will
continue indefinitely. **Existing projects that use `{drake}` can safely continue to use `{drake}`, and there is no need to
retrofit `{targets}`.** New projects should use `{targets}` because it is friendlier and more robust.

# I have some installation problems

The most common cause are missing shared libraries. Please, refer to
[instructions](../index.html#installation-failed) in README.

If those are unsuccessful, feel free to [open a new issue](https://github.com/bioinfocz/scdrake/issues).

# The pipeline is failing for my data

First make sure you have read the [Before you analyse your own data](scdrake.html#before-you-analyse-your-own-data) in the
Quick start vignette (`vignette("scdrake")`).

If the problem persists, feel free to [open a new issue](https://github.com/bioinfocz/scdrake/issues) or start a
[discussion](https://github.com/bioinfocz/scdrake/discussions).

# I want to run the pipeline in parallel mode

```{r, child = "_drake_parallelism.Rmd"}
```

# I want to change the output directory

This can be simply done by changing the appropriate parameters in config files:

- `BASE_OUT_DIR` in `config/{single_sample,integration}/00_main.yaml` is the root directory for all outputs.
- Each stage in each pipeline type has it's own base directory created under `BASE_OUT_DIR`.
  For example, `INPUT_QC_BASE_OUT_DIR` in `config/single_sample/01_input_qc.yaml`.

# I want to use different cache directory

This is controlled by `DRAKE_CACHE_DIR` in `config/pipeline.yaml`.

# I want to load intermediate results

All results of the pipeline (targets) are saved into a `{drake}` cache, and can be simply retrieved using
the `drake::loadd()` or `drake::readd()` functions:

```{r, eval = FALSE}
drake::loadd(name_of_target)
## -- name_of_target can be either quoted (character) or unquoted (symbol)
target <- drake::readd(name_of_target)
```

To know which targets you can load, please, refer to vignettes of individual pipeline stages.

# I want to extend the pipeline

See `vignette("scdrake_extend")`, please.

# I want to manually annotate cells

You can do it using the `CELL_GROUPINGS` or `ADDITIONAL_CELL_DATA_FILE` parameter in
`config/single_sample/02_norm_clustering.yaml` and `config/integration/02_int_clustering.yaml` configs.
Please, refer to `vignette("stage_norm_clustering")` and `vignette("stage_int_clustering")`, respectively,
for description and usage of these parameters.

***

Alternatively, you can reuse a `SingleCellExperiment` object, for example:

```{r, eval = FALSE}
drake::loadd(sce_final_norm_clustering)
umap <- reducedDim(sce_final_norm_clustering, "umap")
cell_types <- dplyr::case_when(
  umap[, 1] > 1 & umap[, 2] < 5 ~ "cell_type_1",
  umap[, 1] > 5 & umap[, 2] < 10 ~ "cell_type_2",
  TRUE ~ "cell_type_3"
)
sce_final_norm_clustering$my_cell_types <- factor(cell_types)
saveRDS(sce_final_norm_clustering, "sce_my_annotation.Rds")
```

We have added a new `colData()` column named `my_cell_types` to the SCE object that divides cells based on their
UMAP coordinates.

Now you need to modify the `INPUT_DATA` parameter in `config/single_sample/01_input_qc.yaml` in order to start the
pipeline from the saved SCE object instead of `cellranger` output:

```yaml
INPUT_DATA:
  type: "sce"
  path: "sce_my_annotation.Rds"
```

The `sce_final_norm_clustering` object is already filtered and normalized, so you should skip those procedures:

```yaml
EMPTY_DROPLETS_ENABLED: False
ENABLE_CELL_FILTERING: False
ENABLE_GENE_FILTERING: False
```

You can also set to skip the normalization step in `config/single_sample/02_norm_clustering.yaml`:

```yaml
NORMALIZATION_TYPE: "none"
```

The `my_cell_types` column can be now used for different purposes, e.g. for cluster markers detection, differential
expression (stage `contrasts`) or visualization.

The same can be done for input data for the integration pipeline. Please, refer to the `INTEGRATION_SOURCES` parameter
in `vignette("stage_integration")`.

# A target is not getting built

It might happen a target (especially RMarkdown one) is not getting built although some changes were introduced.
However, you can either:

- Manually invalidate the target in `{drake}` cache using the `drake::clean()` function and
  it will be built from scratch next time you run the pipeline.
- Use the `DRAKE_REBUILD` parameter in `config/pipeline.yaml` (see `vignette("config_pipeline")`).
