---
title: "Running the pipeline, environment variables"
date: "`r Sys.Date()`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Running the pipeline, environment variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

<script>
$(document).ready(function(){
  $('[data-toggle="tooltip"]').tooltip(); 
});
</script>

***

This vignette expects you have already initialized a new `{scdrake}` project, as described in `vignette("scdrake")`.

# Environment variables

On `{scdrake}` load or
<a data-toggle="tooltip" title="(i.e. when you call some function from a package, e.g. `drake::run_single_sample()`)">attach</a>,
several environment variables are read and corresponding `options()` are set. Those are mainly variables controlling
paths to config directories. See `?scdrake_options` for more details and list of used environment variables.

We can take the `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` environment variable as an example:

- If it is set, the `scdrake_single_sample_config_dir` option is set on `{scdrake}` load or attach.
- It it is not set, a default value `"config/single_sample"` will be used for the option above.
  You can see that the default value is consistent with the file structure of a newly initialized `{scdrake}` project.
- Value of the `scdrake_single_sample_config_dir` option is the default one for many `{scdrake}` functions, e.g.
  `update_single_sample_configs(dir = getOption("scdrake_pipeline_config_dir"), ...)` or
  `run_single_sample(dir = getOption("scdrake_pipeline_config_dir"), ...)`.

# Running the pipeline in current R session

The pipeline can be run in current R session, however, this is considered a bad practice for reproducibility.
The most high-level functions are `run_single_sample()` and `run_integration()` - see description in their help pages
for overview of what they do before pipeline is executed.

Anyway, to replicate what is done inside `run_single_sample()`, you just need this code (assuming you have created
a `{scdrake}` project and your current working directory is in its root):

```{r, eval = FALSE}
update_pipeline_config()
update_single_sample_configs()
cfg_pipeline <- load_pipeline_config()
cfg <- load_single_sample_configs(cfg_pipeline)
plan <- get_single_sample_plan(cfg, cfg_pipeline)
create_single_sample_dirs(cfg)
scdrake_make(plan, cfg_pipeline = cfg_pipeline, cfg_main = cfg_main)
```

Now you have almost full control over what will be executed in the pipeline:

- You can modify `cfg_pipeline` and `cfg`, as these are lists (but with strict rules, see `?scdrake_list`).
  However, you should do modifications with caution, as config parameters are processed by internal functions
  ([config_process_common.R](https://github.com/bioinfocz/scdrake/tree/main/R/config_process_common.R),
  [config_process_single_sample.R](https://github.com/bioinfocz/scdrake/tree/main/R/config_process_single_sample.R),
  [config_process_integration.R](https://github.com/bioinfocz/scdrake/tree/main/R/config_process_integration.R)).
  This might change in the future and these functions will be exposed to package users.
- You can add more targets to `plan`, e.g.

```{r, eval = FALSE}
my_plan <- drake::drake_plan(
  my_target = scater::plotReducedDim(sce_final_norm_clustering, "umap", colour_by = "total")
)
plan <- drake::bind_plans(plan, my_plan)
scdrake_make(plan, cfg_pipeline = cfg_pipeline, cfg_main = cfg_main)
```

Just keep in mind that the list of targets to be made is taken from `cfg_pipeline$DRAKE_TARGETS`, so unless
it is set to `null` (= make all targets in plan) in your `pipeline.yaml` config,
you have to add names of your own targets to be made, i.e.

```{r, eval = FALSE}
cfg_pipeline$DRAKE_TARGETS <- c(cfg_pipeline$DRAKE_TARGETS, "my_target")
scdrake_make(plan, cfg_pipeline = cfg_pipeline, cfg_main = cfg_main)
```

# Running the pipeline reproducibly in a fresh new R session

In `{drake}`, the preferred way to execute pipeline is through `drake::r_make()`, with the `source` parameter set
to an initial script. In `{scdrake}`, there are two such scripts for the single-sample and integration pipeline:
`_drake_single_sample.R` and `_drake_integration.R`, respectively. Loading the config files from proper paths in these
scripts is done via [environment variables](#environment-variables).

```{r, eval = FALSE}
drake::r_make("_drake_single_sample.R")
drake::r_make("_drake_integration.R")
```

All important details are already given in *Running the pipeline reproducibly* section in `vignette("scdrake")`.
