---
title: "Advanced topics"
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Advanced topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette should serve as a supplement to the Get Started vignette (`vignette("scdrake")`) and
expects you have finished the initial three steps.

Other "hot" topics and questions can be found in `vignette("scdrake_faq)`

***

## Retrieving intermediate results (targets)

In `{drake}`'s terminology, a pipeline is called *plan*, and is composed of *targets*. When a target is finished,
its value (object) is saved to cache (the directory `.drake` by default). The cache has two main purposes:

- If a target is not changed, its value is loaded from the cache. *Change* involves e.g. target's definition (code) or
  change in upstream targets on which the target depends. This way `{drake}` is able to skip computation of finished
  targets and greatly enhance the runtime. More details [here](https://books.ropensci.org/drake/triggers.html).
- Users also have access to the cache, and so you can load any finished target into your R session.

Users can access the cache via two `{drake}`'s functions:

- `drake::loadd()` loads target's value to the current session as a variable of the same name.
- `drake::readd()` returns target's value (so it can be assigned to variable).

Let's try it and load the filtered `SingleCellExperiment` object:

```{r}
drake::loadd(sce_final_input_qc)
```

Value of the target `sce_final_input_qc` was loaded as a variable of the same name to your current R session
(or more precisely, to the global environment).

Similarly, we can load this target to a variable of our choice:

```{r}
sce <- drake::readd(sce_final_input_qc)
```

And work further with the loaded object, e.g.

```{r}
scater::plotExpression(sce, "NOC2L", exprs_values = "counts", swap_rownames = "SYMBOL")
```

***

## How to dig into `scdrake` plans?

For a more schematic overview of pipelines and stages see `vignette("pipeline_overview")`, where are also
[diagrams](https://github.com/bioinfocz/scdrake/blob/main/diagrams/README.md).

Advanced users might be interested in looking into source code of `{scdrake}`'s [plans](https://github.com/bioinfocz/scdrake/tree/main/R)
(files named `plans_*.R`).

***

## Running the pipeline in parallel mode

```{r, eval = TRUE, child = "_drake_parallelism.Rmd"}
```

***

## Using an alternative storage format (`qs`)

By default, R's `Rds` format is used (see `?saveRds`) to save intermediate results to `{drake}`s cache,
but instead, we recommend to use `DRAKE_FORMAT: "qs"` (see <https://github.com/traversc/qs>) in `config/pipeline.yaml`
that offers better performance, but sometimes doesn't work correctly (`drake` throws untraceable errors).

See <https://books.ropensci.org/drake/plans.html#special-data-formats-for-targets> for more details.

***

## Extending the pipeline

See `vignette("scdrake_extend")`, please.

***

## Ligand receptor interactions in Python

See `vignette("interaction_analysis")`, please.

***
