---
title: "Using the Docker image"
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Using the Docker image}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, child = "_version.Rmd"}
```

***

[Docker](https://www.docker.com/) ([wiki](https://en.wikipedia.org/wiki/Docker_(software))) is a software for running
isolated *containers* which are based on *images*. You can think of them as small, isolated operating systems.

The [Bioconductor's page for Docker](https://bioconductor.org/help/docker/) provides itself a nice introduction to
Docker and its usage for R, and we recommend to read it through.
In this vignette, we will mainly give practical examples related to `{scdrake}`.

***

# Obtaining the Docker image

A Docker image based on the [official Bioconductor image](https://bioconductor.org/help/docker/)
(version `r LATEST_STABLE_VERSION_DOCKER_BIOC`) is available. This is the most handy and reproducible way how to use
`{scdrake}` as all the dependencies are already installed and their version is fixed.
In addition, the parent Bioconductor image comes bundled with RStudio Server.

You can pull the Docker image with the latest stable `{scdrake}` version using

```{r docker_stable, include = FALSE}
out <- scdrake::format_shell_command(glue::glue("docker pull bioinfocz/scdrake:{LATEST_STABLE_VERSION}-bioc{LATEST_STABLE_VERSION_DOCKER_BIOC}"))
```

`r knitr::knit(text = out)`

or list available versions in [our Docker Hub repository](https://hub.docker.com/r/bioinfocz/scdrake/tags).

For the latest development version use

```{r docker_latest, include = FALSE}
out <- scdrake::format_shell_command(glue::glue("docker pull bioinfocz/scdrake:latest-bioc{LATEST_STABLE_VERSION_DOCKER_BIOC}"))
```

`r knitr::knit(text = out)`

***

# Running the image

The `{scdrake}`'s image is based on the [Bioconductor's image](https://bioconductor.org/help/docker/)
which is in turn based on the [Rocker Project's](https://rocker-project.org/images/versioned/rstudio.html)
image (`rocker/rstudio`). Thanks to it, the `{scdrake}`'s image comes bundled with RStudio Server.

Docker allows to mount local (host) directories to containers. We recommend to create a local directory in
which individual `{scdrake}` projects (and other files, if needed) will lie. This way you won't lose data when
a container is destroyed. Let's create such shared directory in your home directory and switch to it:

```bash
mkdir ~/scdrake_projects
cd ~/scdrake_projects
```

## Running with RStudio Server

[RStudio Server](https://posit.co/products/open-source/rstudio-server/) is a comfortable IDE for R.
You can run the image including an RStudio instance using:

```{r docker_run_rstudio, include = FALSE}
out <- scdrake::format_shell_command(c(
  "docker run -d",
  "-v $(pwd):/home/rstudio/scdrake_projects",
  "-p 8787:8787",
  "-e USERID=$(id -u)",
  "-e GROUPID=$(id -g)",
  "-e PASSWORD=1234",
  glue::glue("scdrake:{LATEST_STABLE_VERSION}-bioc{LATEST_STABLE_VERSION_DOCKER_BIOC}")
))
out2 <- glue::glue("`scdrake:{LATEST_STABLE_VERSION}-bioc{LATEST_STABLE_VERSION_DOCKER_BIOC}`")
```

`r knitr::knit(text = out)`

Let's decompose the command above:

- `docker run`: run image as a container.
- `-d`: run in detached ("background") mode.
- `-v $(pwd):/home/rstudio/scdrake_projects`: mount the current working directory on the host machine as
  `/home/rstudio/scdrake_projects` inside the container.
- `-p 8787:8787`: expose port `8787` in container as port `8787` on localhost.
- `-e USERID=$(id -u) -e GROUPID=$(id -g)`: change ownership of `/home/rstudio` in the container to the ID and group ID
  of the current host user. This is important as it sustains the correct file ownership, i.e. when you write to
  `/home/rstudio/scdrake_projects` from within the container, you actually write to the host filesystem and you do so
  as the proper user/group.
- `-e PASSWORD=1234`: password for RStudio Server (and the username is `rstudio`).
- `r out2`:
  repository and tag of the image to be run.

The `-e VAR=VALUE` arguments are used to set environment variables inside the container. You can check those for the
parent `rocker/rstudio` image [here](https://rocker-project.org/images/versioned/rstudio.html#environment-variables).

An image can have a default command which is executed on container run. For our image, it's a script `/init` which
starts RStudio Server.

Now you can check that the container is running:

```bash
docker ps
```

You should see something as

```
CONTAINER ID   IMAGE                            COMMAND   CREATED        STATUS        PORTS                                       NAMES
d47b4d265052   scdrake:1.4.0-bioc3.15           "/init"   24 hours ago   Up 24 hours   0.0.0.0:8787->8787/tcp, :::8787->8787/tcp   condescending_payne
```

Values in the `CONTAINER ID` and `NAMES` columns can be then used to reference the running container.

You can see that we have exposed port `8787`. What it is useful for? Well, it allows us to connect to RStudio Server,
which is running on port 8787 inside the container. Now you can just open your browser, navigate to `localhost:8787`
and login as `rstudio` with password `1234`.

TIP: if you are using a remote server which you can SSH into and it allows SSH tunneling, then you can forward the
exposed port to your host:

```bash
ssh -NL 8787:localhost:8787 user@server
```

Now you can start using `{scdrake}` from the RStudio, or read below for alternative ways.

## Run the image without RStudio Server

If you just want to run the image detached without RStudio Server, use

```{r docker_run_detached, include = FALSE}
out <- scdrake::format_shell_command(c(
  "docker run -d",
  "-u rstudio",
  "-v $(pwd):/home/rstudio/scdrake_projects",
  glue::glue("scdrake:{LATEST_STABLE_VERSION}-bioc{LATEST_STABLE_VERSION_DOCKER_BIOC}"),
  "sleep infinity"
))
```

`r knitr::knit(text = out)`

***

# Issuing commands in the running container

You can start `bash` or `R` process inside the container and attach to it using

```bash
docker exec -it -u rstudio <CONTAINER ID or NAME> bash
docker exec -it -u rstudio <CONTAINER ID or NAME> R
```

- `docker exec`: executes a command in a **running** container.
- `-i`: interactive.
- `-d`: allocate a pseudo-TTY.

***

# Problems with filesystem permissions {.tabset}

In the first example, when we run the image including RStudio Server, we specified
`-e USERID=$(id -u) -e GROUPID=$(id -g)`, and so ownership of `/home/rstudio` was changed to match the host user\*.
Then when you write to the shared directory `/home/rstudio/scdrake_projects`, you are basically doing that as the host user.

*<sup>\* This is actually not a feature of Docker, but the `rocker/rstudio` image which uses those environment
variables to run the `/rocker_scripts/init_userconf.sh` which manages the ownership.</sup>*

By default, `rstudio` user in the container has ID of 1000 which is commonly used for the default user in most
Linux distributions. You can check it yourself on your host by executing `id -u`. If it is 1000, then you can safely
skip the instructions below.

***

So when you don't want to run RStudio Server and your host user ID is not 1000, you have to take care of ownership
by yourself. There are several ways how to accomplish that. You can also read about this problem
[here](https://www.fullstaq.com/knowledge-hub/blogs/docker-and-the-host-filesystem-owner-matching-problem) or
[here](https://vsupalov.com/docker-shared-permissions/).

## Modify the user and group ID in the container

The `rocker/rstudio` parent image already comes with a script which can add or modify user information
(it is actually run when you start the container including RStudio Server).

```bash
docker exec \
  -e USERID=$(id -u) -e GROUPID=$(id -g) \
  <CONTAINER ID or NAME> \
  bash /rocker_scripts/init_userconf.sh
```

Note that this just changes ID and group of the `rstudio` user, so later don't forget to run commands as this user,
i.e. `docker exec -u rstudio ...`

## "Chowning" (tedious)

If you have root privileges on the host, you can just change ownership each time you write to the shared directory
from the container:

```bash
sudo chown -R $(id -u):$(id -g) scdrake_projects
```

***

# Issuing `{scdrake}` commands through its command line interface (CLI)

We now assume that the `{scdrake}` container is running and the shared directory is mounted in
`/home/rstudio/scdrake_projects`. The `{scdrake}` commands can be easily issued through its CLI
(see `vignette("scdrake_cli")`):

```bash
docker exec -it -u rstudio <CONTAINER ID or NAME> \
  scdrake -h
```

The usual workflow can be similar to the following:

1. Initialize a new project:

```bash
docker exec -it -u rstudio -w /home/rstudio/scdrake_projects <CONTAINER ID or NAME> \
  scdrake -d my_first_project init-project
```

2. Modify the configs on your host in `my_first_project/config/`

3. Run the pipeline:

```bash
docker exec -it -u rstudio -w /home/rstudio/scdrake_projects/my_first_project <CONTAINER ID or NAME> \
  scdrake --pipeline-type single_sample run
```

***

# Create a command alias

To reduce typing/copy-pasting a bit, you can create a command alias in your shell, here for `bash`:

```bash
alias scdrake_docker="docker exec -it -u rstudio <CONTAINER ID or NAME> scdrake"
```

And then use this alias as e.g.

```bash
scdrake_docker -d /home/rstudio/scdrake_projects/my_first_project --pipeline-type single_sample run
```

Notice that we are using the `-d` parameter as the alias doesn't contain the `-w` parameter which instructs Docker
to execute the command in a working directory.

Finally, you can put the alias permanently inside your `~/.bashrc` file. Just don't forget to change
`<CONTAINER ID or NAME>` when you start a new container :)
