---
title: "scdrake config internals"
description: "How YAML config files are used in `scdrake`"
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{scdrake config internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

***

# Introduction

`{scdrake}` configs are stored in YAML files (in the `config/` directory by default), which are parsed in R to lists
and used to build `{drake}` plans.
YAML ([quick reference](https://docs.ansible.com/ansible/latest/reference_appendices/YAMLSyntax.html),
cheatsheets [here](https://lzone.de/cheat-sheet/YAML) and [here](https://learnxinyminutes.com/docs/yaml/))
is a human-readable format and uses indentation (spaces) for nesting of parameters.

`{scdrake}` is using a concept of default and local configs. Default configs are bundled with the package and copied
during project initialization, update, and pipeline run (e.g. in `run_single_sample()` or in `_drake_single_sample.R`).
Local configs are, as their name suggests, purposed to make modifications to default configs
(see the section below for how it is actually done).
Default and local configs are named `*.default.yaml` and `*.yaml`, respectively.

Config files are separated according to general parameters and each pipeline's main stage
(quality control, normalization, etc.), but for consistency are always read all at once
(by default if you run e.g. `run_single_sample()` or `run_single_sample_r()`).

All paths in configs must be relative to project root, or absolute (not recommended), unless otherwise stated.

# Updating (merging) configs

It may happen that new parameters are added to default configs. In that case, those new parameters need to be appended
to local configs, while preserving the current local parameters. Calling this procedure "update" is a little bit misleading,
as we actually overwrite default config by local one and append new parameters from default config to it.
"Merging" should be a better word, as we *merge* default config with local one (and the order does matter).

Configs are merged recursively by parameter names. It is necessary to realize that YAML format are nested dictionaries
(or named lists in the context of R). That is, this YAML

```yaml
# Default config.
FOO: 1
BAR: "baz"
FOO_LIST: [1, "hello"]
FOO_NAMED_LIST:
  FOO_2: 2
  BAR_2: "zab"
```

is in R parsed to the list:

```r
list(FOO = 1L, BAR = "baz", FOO_LIST = list(1L, "hello"), FOO_NAMED_LIST = list(FOO_2 = 2L, BAR_2 = "zab"))
```

So, for example, given that the YAML above is our default config, we want to merge it with the local config

```yaml
# Local config.
BAR: "zab"
FOO_LIST: [4, 5, 6]
FOO_NAMED_LIST:
  FOO_2: 3
  BAR_3: 5
```

resulting in

```yaml
# Updated local config.
FOO: 1
BAR: "zab"
FOO_LIST: [4, 5, 6]
FOO_NAMED_LIST:
  FOO_2: 3
  BAR_2: "zab"
  BAR_3: 5
```

- `FOO` is not updated.
- `BAR` and `FOO_LIST` are overwritten by the local config.
- `FOO_NAMED_LIST`: `FOO_2` is updated, `BAR_3` is added, but `BAR_2` is still present, although it is not defined
  in the local config!

## Merging of nested named lists

To overcome the problem with `FOO_NAMED_LIST`, parameters with such structure are specified as a named list wrapped
by an unnamed list:

```yaml
FOO_NAMED_LIST:
  - FOO_2: 2
    BAR_2: "zab"
```

**Note the beginning `-` creating the unnamed list and indentation of values of the named list.**
This YAML is parsed in R to

```r
list(FOO_NAMED_LIST = list(list(FOO_2 = 2L, BAR_2 = "zab")))
```

If we modify the example of local config above to

```yaml
FOO_NAMED_LIST:
  - FOO_2: 3
    BAR_3: 5
```

then the whole `FOO_NAMED_LIST` parameter will be replaced during a config merge, and that is the desired behaviour.

## Using structure of a default config

Also, by default, a structure (parameter positions, comments) of local config is preserved during update.
However, it is possible to use the structure of a default config - see `?update_config` for more details.

# Using R code in parameters

A special type of parameter starting with `!code` can be used to evaluate a value as R code:

```yaml
EXAMPLE: !code 1:3
```

First, non-code parameters are loaded to a separate environment and then code parameters are evaluated in the
context of this environment. This means you can use other config parameters as R variables inside code parameters:

```yaml
FOO: 1
BAR: !code FOO + 1
```

In addition, in stage configs (e.g. `02_norm_clustering`) you can also use parameters (variables) from `pipeline.yaml`
and `00_main.yaml` configs:

```yaml
EXAMPLE: !code glue("{PROJECT_NAME}_{INSTITUTE}")
```

See `?load_config` for more details.

# The `yq` tool

Internally, the [yq](https://github.com/mikefarah/yq) tool (version 3) is used for merging of YAML files.
It is a command line utility whose binary needs to be downloaded. This is done automatically during initialization of
a new project, or you can do it manually through `download_yq()`.

On `{scdrake}` load or attach, `SCDRAKE_YQ_BINARY` environment variable is read - if not set,
a value from `Sys.which("yq")` is used (this function searches in `PATH` environment variable).
Then `scdrake_yq_binary` option is set, and is used as default value to config-updating functions (see `?update_config`).

You can also look at `?check_yq` for details on how `PATH` variable is treated in terminal vs. RStudio.

# `scdrake_list`

The base R's extracting operators (`$`, `[`, `[[`) are very benevolent for non-existing elements in `list()`
for which return `NULL`. This is not a desired behaviour for config variables that must have a value or explicit
`NULL`. Returning `NULL` when the value was actually not loaded from a config file can lead to unpredictable results.

Thus, for storing config variables, `{scdrake}` is using a modified `list()` called `scdrake_list()`
which is using strict extracting operators. It behaves like a normal list:

```{r, message = FALSE}
cfg <- scdrake::scdrake_list(list(var_1 = 1, var_2 = 2))
cfg$var_1
cfg[["var_2"]]
cfg["var_1"]
cfg[c("var_1", "var_2")]
```

But extracting non-existing elements throws error:

```{r, error = TRUE}
cfg$var_3
cfg[["var_3"]]
cfg["var_3"]
```

For `[` and `[[`, this control can be turned off by `check = FALSE` parameter:

```{r}
cfg[["var_3", check = FALSE]]
cfg["var_3", check = FALSE]
```

Also, in this case `[` is more consistent: it returns a valid named list, unlike the normal list, which sets
`NA_character_` as names of the non-existing elements.
