---
title: "Input and quality control stage"
description: "01_input_qc stage of the single-sample pipeline."
date: "`r Sys.Date()`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Input and quality control stage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

***

For a general overview of this single-sample pipeline stage see `vignette("pipeline_overview")`.

Config for this stage is stored in `01_input_qc.yaml` file.
Directory with this file is read from `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` environment variable upon `{scdrake}` load or
attach, and saved as `scdrake_single_sample_config_dir` option. This option is used as the default argument value in
several `{scdrake}` functions.

# Config parameters

## Removal of empty droplets

See `?DropletUtils::emptyDrops` for more details.

***

`EMPTY_DROPLETS_LOWER: 100`

**Type:** positive integer scalar

A lower bound on the total UMI count at or below which all barcodes are assumed to correspond to empty droplets.

***

`EMPTY_DROPLETS_FDR_THRESHOLD: 0.01`

**Type:** numeric scalar <0; 1>

A threshold for FDR adjusted p-values for null hypothesis that barcode comes from ambient environment.
In other words, probability that a droplet was empty and contained ambient RNA.

## Cell filtering

Two types of cell filtering are performed: *dataset-sensitive* (using MAD threshold),
and *custom* (using custom thresholds). See `vignette("pipeline_overview")` for more details.

In both filtering types, violation of only one metric threshold leads to removal of a cell.

***

`SAVE_DATASET_SENSITIVE_FILTERING: True`

**Type:** logical scalar

If `True`, proceed to other stages with dataset filtered by dataset-sensitive filtering (target `sce_qc_filter_genes`),
otherwise with dataset filtered by custom filtering (target `sce_custom_filter_genes`).

### Dataset-sensitive cell filtering

`MAD_THRESHOLD: 3`

**Type:** positive numeric scalar

A threshold for maximum MAD (median absolute deviation) for QC cell metrics:

- Number of UMI (lower tail).
- Number of detected genes = non-zero UMI count (lower tail).
- Percentage of expressed mitochondrial genes = $\frac {UMI_{mitochondrial}} {UMI_{sum}} * 100$ (upper tail).

"Lower tail" means cells having a metric value less than `-MAD_THRESHOLD` MAD will be discarded.
For "upper tail", it is `+MAD_THRESHOLD` MAD.

Violation of only one metric threshold leads to removal of a cell.

### Custom cell filtering

`MIN_UMI_CF: 1000`

**Type:** positive integer scalar

A threshold for minimum number of UMI per cell, i.e. cells with UMI less than `MIN_UMI_CF` will be removed.

***

`MAX_UMI_CF: 50000`

**Type:** positive integer scalar

A threshold for maximum number of UMI per cell, i.e. cells with UMI greater than `MAX_UMI_CF` will be removed.

***

`MIN_FEATURES: 1000`

**Type:** positive integer scalar

A threshold for minimum number of features (genes) detected per cell,
i.e. cells with detected features less than `MIN_FEATURES` will be removed.

***

`MAX_MITO_RATIO: 0.2`

**Type:** numeric scalar <0; 1>

A threshold for maximum ratio of expressed mitochondrial genes per cell,
i.e. cells with mitochondrial genes detected in more than (`MAX_MITO_RATIO` * 100)% of all genes will be removed.

### Gene filtering

Gene filtering thresholds are applied in both types of cell filtering. The filter is computed by the
`get_gene_filter()` function as following:

```{r, eval = FALSE}
num_cells <- min_ratio_cells * ncol(sce)
is_expressed <- rowSums(counts(sce) >= min_umi) >= num_cells
```

A gene is considered expressed when number of its UMIs across all cells is greater than `min_umi` and
at the same time it is expressed in at least `min_ratio_cells` ratio of cells.

***

```yaml
MITO_REGEX: "^MT-"
RIBO_REGEX: "^RP[SL]"
```

**Type:** character scalar

Regexes used to match and count occurences of mitochondrial and ribosomal features.
The latter is currently not used for cell filtering.

***

`MIN_UMI: 1`

**Type:** positive integer scalar

A threshold for minimum number of UMI per cell, i.e. genes with UMI < `MIN_UMI` will be removed.

***

`MIN_RATIO_CELLS: 0.01`

**Type:** numeric scalar <0; 1>

A minimum ratio of cells expressing a gene,
i.e. genes expressed in less than (`MIN_RATIO_CELLS` * 100)% of cells will be removed.

## Input files

`INPUT_10X_DIR: "/path/to/dir"`

**Type:** character scalar

A path to directory (could be relative to the project root) containing output files from `cellranger count` command,
that is, `barcodes.tsv`, `features.tsv`, and `matrix.mtx` (can be also gzipped with `.gz` extension).
Passed to `DropletUtils::read10xCounts()`.

You can download example PBMC data from 10x Genomics using the `download_pbmc1k()` and `download_pbmc1k()` functions.

***

`INPUT_QC_REPORT_RMD_FILE: "Rmd/single_sample/01_input_qc.Rmd"`

**Type:** character scalar

A path to RMarkdown file used for HTML report of this pipeline stage.

## Output files

`INPUT_QC_BASE_OUT_DIR: "01_input_qc"`

**Type:** character scalar

A path to base output directory for this stage. It will be created under `BASE_OUT_DIR` specified in `00_main.yaml` config.

***

`INPUT_QC_REPORT_HTML_FILE: "01_input_qc.html"`

**Type:** character scalar

A name of HTML report file for this stage. Created in `INPUT_QC_BASE_OUT_DIR`. Subdirectories are not allowed.

## HTML output parameters

```yaml
INPUT_QC_KNITR_MESSAGE: False
INPUT_QC_KNITR_WARNING: False
INPUT_QC_KNITR_ECHO: False
```

**Type:** logical scalar

These are passed to `knitr::opts_chunk()` and used for rendering of stage's HTML report.

# Targets

Here you can find description of the most important targets for this stage.
However, for a full overview, you have to inspect the
[source code](https://github.com/bioinfocz/scdrake/blob/main/R/plans_single_sample.R) of the
`get_input_qc_subplan()` function.

## `SingleCellExperiment` objects

`sce_raw`: an untouched SCE object as loaded from `DropletUtils::read10xCounts()`.

***

`sce_valid_cells`: a SCE object with empty droplets removed.

***

`sce_unfiltered`: `sce_valid_cells` with added columns:

- Columns from the `cell_qc` target containing DataFrame with cell QC metrics.
- `discard_qc`: dataset-sensitive filter.
- `discard_custom`: custom filter.

***

`sce_qc_filter`, `sce_qc_filter_genes`: a SCE object with cells filtered by dataset-sensitive filtering.
The latter with filtered genes.

`sce_custom_filter`, `sce_custom_filter_genes`: a SCE object with cells filtered by custom filtering.
The latter with filtered genes.

***

`sce_selected`: a SCE object selected from either
`sce_qc_filter_genes` or `sce_custom_filter_genes` - depends on the `SAVE_DATASET_SENSITIVE_FILTERING` parameter.

`sce_final_input_qc`: `sce_selected` with added gene annotation (`gene_annotation` target) to `rowData()`.
This is the final SCE object which will be used in the next stage `norm_clustering`.

## Cell filters

`cell_qc`: a DataFrame with cell QC metrics.

***

The following list of filters are computed by `scater::isOutlier()` from `cell_qc` columns,
which are shown for individual filters (e.g. `qc_lib`: `total` column).

`qc_filters`: a list of named lists with logical values for dataset-sensitive cell filtering:

- `qc_lib` (`total`): low number of UMI (lower tail).
- `qc_nexprs` (`detected`): low number of detected genes = non-zero UMI count (lower tail).
- `qc_mito` (`subsets_mito_percent`): high percentage of expressed mitochondrial genes expression (upper tail).

`custom_filters`: a list of named lists with logical values for custom cell filtering:

- `low_count` (`total`): low number of UMIs.
- `high_count` (`total`): high number of UMIs.
- `low_expression` (`detected`): low number of detected genes.
- `high_mito` (`subsets_mito_percent`): high percentage of expressed mitochondrial genes.

`qc_filter`, `custom_filter`: a summarized filter (by "or" operation) of `qc_filters` and `custom_filters`, respectively.

## Gene filters

`sce_qc_gene_filter`, `sce_custom_gene_filter`: gene filters (logical vectors) obtained from `get_gene_filter()` for
SCE objects with cells filtered by dataset-sensitive and custom filter, respectively.

## Plots

`sce_unfiltered_plotlist`: a list of various violin plots for cell QC metrics (`{ggplot2}` objects), generated using
`sce_unfiltered` target. Cells are colored by `discard_qc` column - blue points pass the dataset-sensitive filter,
while the orange ones do not.

For example, the following code is used to plot total UMI counts:

```{r, eval = FALSE}
plot_colData(
  sce_unfiltered,
  y = "total",
  colour_by = "discard_qc",
  title = "Total count",
  scale_y = ggplot2::scale_y_log10()
)
```

***

`sce_qc_filter_genes_plotlist`, `sce_custom_filter_genes_plotlist`: similar to `sce_unfiltered_plotlist`,
but in each of them, cells are colored by the other filter. That is, cells in `sce_qc_filter_genes_plotlist` are
colored by `discard_custom` column, and thus, showing which cells would be discarded in custom-filtered dataset.
Vice versa for `sce_custom_filter_genes_plotlist`.

## Other targets

`config_input_qc`: a list holding parameters for this stage.

***

`barcode_ranks`: a DataFrame with barcode ranks for
[knee plot](https://bioconductor.org/books/release/OSCA/droplet-processing.html),
computed by `DropletUtils::barcodeRanks()` on UMI counts from `sce_raw` target.

`empty_droplets`: a DataFrame of statistics for empty droplets, computed `DropletUtils::emptyDrops()` on UMI counts
from `sce_raw` target.

***

`sce_history`: a tibble with cell and gene "history", showing their numbers in unfiltered and filtered data.

`sce_history_plot`: a `{ggplot2}` object summarizing information from `sce_history` target.

***

`gene_annotation`: a dataframe with gene annotation, processed such that:

- If a single ENSEMBL ID has multiple symbols, gene descriptions, or ENTREZ IDs, they are collapsed by comma (`,`).
- ENSEMBL ID is used as a symbol for ENSEMBL IDs with unknown symbols.
- ENSEMBL ID is appended to symbols having multiple ENSEMBL IDs.
  - For example, *TBCE* has both ENSG00000285053 and ENSG00000284770 ENSEMBL IDs assigned ->
    its symbol is changed to TBCE_ENSG00000285053 and TBCE_ENSG00000284770.

`gene_annotation` can be also retrieved from `rowData(sce_final_input_qc)`.
