---
title: "Cluster markers stage (`cluster_markers`)"
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Cluster markers stage (`cluster_markers`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

***

## {.tabset}

### Overview

A stage for calculation, visualization and reporting of cell cluster markers ("global markers").
This stage is common to both single-sample and integration pipelines.

`r emoji::emoji("gear")` Config files: `config/single_sample/cluster_markers.yaml`, `config/integration/cluster_markers.yaml`

`r emoji::emoji("clipboard")` HTML report target (in `config/pipeline.yaml`):
`DRAKE_TARGETS: ["report_cluster_markers"]`

[`r emoji::emoji("scroll")` Example report for PBMC 1k data](https://onco.img.cas.cz/novotnyj/scdrake/single_sample/pbmc1k/cluster_markers/cluster_markers.html)
([used config](https://github.com/bioinfocz/scdrake/blob/main/tests/testthat/run_pipeline_config_patches/single_sample_pbmc1k_full/cluster_markers.default.yaml))

[`r emoji::emoji("scroll")` Example report for integrated data](https://onco.img.cas.cz/novotnyj/scdrake/integration/cluster_markers/cluster_markers.html)
([used config](https://github.com/bioinfocz/scdrake/blob/main/tests/testthat/run_pipeline_config_patches/integration/cluster_markers.default.yaml))

`r emoji::emoji("ladder")` Structure

- [Calculation of cluster markers](https://bioconductor.org/books/3.15/OSCA.basic/marker-detection.html):
  any categorical grouping of cells can be used, including those manually defined based on the existing groups from 
  clustering (but not only). See below for more details about marker calculation.
- Generation of heatmaps and plots. To reduce the size of heatmaps and number of plots, you can control a number of top
  N markers taken by a chosen metric (FDR, LFC, etc.). Only these top N markers will be displayed in heatmaps, and plots
  made for them. This can be specified for each cell grouping and test type.
  - Marker plots are composed of:
    - Two plots of a dimred: one colored by a clustering, and second colored by expression of a marker (feature plot).
    - A plot of summarized marker expression: proportion of cells expressing a marker, colored by average expression.
    - A violin plot: expression of a marker in each cluster.
- Generation of marker tables for each cell grouping and each of its levels.

***

#### Marker calculation and interpretation

Taken from [OSCA](https://bioconductor.org/books/3.15/OSCA.basic/marker-detection.html), slightly modified:

> To interpret our clustering results, we identify the genes that drive separation between clusters.
> These marker genes allow us to assign biological meaning to each cluster based on their functional annotation.
> In the most obvious case, the marker genes for each cluster are a priori associated with particular cell types,
> allowing us to treat the clustering as a proxy for cell type identity. The same principle can be applied to discover
> more subtle differences between clusters (e.g., changes in activation or differentiation state) based on the behavior
> of genes in the affected pathways.

> Identification of marker genes is usually based around the retrospective detection of differential expression
> between clusters. Genes that are more strongly DE are more likely to have caused separate clustering of cells
> in the first place. Several different statistical tests are available to quantify the differences in expression
> profiles, and different approaches can be used to consolidate test results into a single ranking of genes for
> each cluster. These choices parametrize the theoretical differences between the various marker detection strategies.

##### Used statistical tests

For each cell grouping and gene, three distinct, adjustable statistical tests are computed through `scran::findMarkers()`:

- t-test (`scran::pairwiseTTests()`): test for difference in mean expression.
- [Wilcoxon signed-rank test](https://en.wikipedia.org/wiki/Wilcoxon_signed-rank_test)
  (`scran::pairwiseWilcox()`): test for difference in median expression (roughly speaking).
- Binomial test (`scran::pairwiseBinom()`):
  test the null hypothesis that the proportion of cells expressing a gene is the same between groups.

These tests are performed in a
[pairwise fashion](https://bioconductor.org/books/3.15/OSCA.basic/marker-detection.html#scoring-markers-by-pairwise-comparisons)
within each grouping, i.e. each level of the grouping is tested against each other.

Example: three pairwise tests of each type are computed for k-means clustering with $k = 3$:

- Cluster 1 vs. Cluster 2
- Cluster 1 vs. Cluster 3
- Cluster 2 vs. Cluster 3

The overall p-values, false discovery rate (FDR) and effect sizes
(log2 fold change (LFC) for t- and binomial test, and Area Under Curve for Wilcoxon test)
are calculated for the cell grouping from its all pairwise tests (`scran::combineMarkers()`).

##### Narrowing down the number of cluster markers

A more or less stringent approach can be used to narrow down the number of cluster markers,
based on results from individual pairwise tests. This will also affect the combined p-value and FDR:

- [Looking for any differences](https://bioconductor.org/books/3.12/OSCA/marker-detection.html#looking-for-any-differences):
  for each level (cluster) of a grouping, take top markers by p-value from each pairwise test. This will be shown
  as `Top` column in marker table. Example: the set of all genes with `Top` values of 1 contains the gene with
  the lowest p-value from each comparison.
- [Finding cluster-specific markers](https://bioconductor.org/books/3.12/OSCA/marker-detection.html#finding-cluster-specific-markers):
  a more stringent approach only considering genes that are differentially expressed in all pairwise comparisons
  involving the cluster of interest. For this purpose is used an intersection-union test where the combined
  p-value for each gene is the maximum of the p-values from all pairwise comparisons.
  A gene will only achieve a low combined p-value if it is strongly DE in all comparisons to other clusters.
- [Balancing stringency and generality](https://bioconductor.org/books/3.12/OSCA/marker-detection.html#balancing-stringency-and-generality):
  the Holm-Bonferroni correction is applied on pairwise p-values and the middle-most value is taken as the combined
  p-value. This effectively tests the global null hypothesis that at least 50% of the individual pairwise comparisons
  exhibit no differential expression.

This can be controlled via the `PVAL_TYPE` parameter.

##### Handling blocking factors

[Blocking factors](https://bioconductor.org/books/3.12/OSCA/marker-detection.html#marker-batch)
(e.g. batch effect, sex differences, cell cycle phases) can be handled by further nesting.
For that, each pairwise test is performed separately in each level of the blocking factor.
Then, p-values from individual levels' tests are combined, and the final combined p-values are obtained by
the method of choice (see above).

This can be controlled via the `BLOCK_COLUMN` parameter.

***

### Config parameters

Cluster markers config is stored in the `config/single_sample/cluster_markers.yaml` and `config/integration/cluster_markers.yaml`
files (the location of this file is different for the single-sample and integration pipelines).
As for the [pipeline config](config_pipeline.html), directory with this file is read from environment variables:

- `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` for the single-sample pipeline
- `SCDRAKE_INTEGRATION_CONFIG_DIR` for the integration pipeline

Options named in lowercase are set upon `{scdrake}` load or attach. Then the actual directory used depends on
whether you run `run_single_sample_r()` or `run_integration_r()`.

***

#### Cluster markers

```yaml
CLUSTER_MARKERS_SOURCES_DEFAULTS:
  COMMON_PARAMS:
    PLOT_DIMREDS: ["umap"]
    BLOCK_COLUMN: null
  PARAMS_T:
    LFC_DIRECTION: "up"
    LFC_TEST: 0
    PVAL_TYPE: "any"
    MIN_PROP: null
    STD_LFC: False
    TOP_N_HEATMAP: 10
    TOP_N_WT_HEATMAP: "top"
    TOP_N_PLOT: 5
    TOP_N_WT_PLOT: "top"
  PARAMS_WILCOX:
    LFC_DIRECTION: "up"
    LFC_TEST: 0
    PVAL_TYPE: "any"
    MIN_PROP: null
    STD_LFC: null
    TOP_N_HEATMAP: 10
    TOP_N_WT_HEATMAP: "top"
    TOP_N_PLOT: 5
    TOP_N_WT_PLOT: "top"
  PARAMS_BINOM:
    LFC_DIRECTION: "up"
    LFC_TEST: 0
    PVAL_TYPE: "any"
    MIN_PROP: null
    STD_LFC: null
    TOP_N_HEATMAP: 10
    TOP_N_WT_HEATMAP: "top"
    TOP_N_PLOT: 5
    TOP_N_WT_PLOT: "top"
```

**Type:** named list (dictionary) of named lists

Default parameters for computation and reporting of cluster markers. These can be overriden for each of the cluster
markers source in `CLUSTER_MARKERS_SOURCES` (see below).

- `COMMON_PARAMS`: used for all statistical tests and their results.
  - `PLOT_DIMREDS` (character vector: `"pca"`, `"tsne"`, `"umap"`): which dimensionality reductions to use for marker plots.
  - `BLOCK_COLUMN` (character scalar or `null`): which column to use as the blocking factor.
    Examples: `phase` (cell cycle phase), `batch` (sample of origin - integration pipeline).
- `PARAMS_T`, `PARAMS_WILCOX`, `PARAMS_BINOM`: parameters for individual test types.
  - Computation parameters (passed to `scran::findMarkers()`, `scran::pairwiseTTests()`, `scran::pairwiseWilcox()`,
    or `scran::pairwiseBinom()`):
    - `LFC_DIRECTION` (character scalar: `"any"` | `"up"` | `"down"`):
      the direction of log-fold changes to be considered in the alternative hypothesis.
    - `LFC_TEST` (positive numeric scalar): the log-fold change threshold to be tested against.
    - `PVAL_TYPE` (character scalar: `"any"` | `"some"` | `"all"`):
      how p-values are to be combined across pairwise comparisons for a given group/cluster.
      If `"any"`, `top` column will be present in cluster markers tables.
      By this parameter you can control the stringency of marker detection.
    - `MIN_PROP` (numeric scalar: `<0; 1>`, or `null`):
      the minimum proportion of significant comparisons per gene.
      Defaults to `0.5` when `PVAL_TYPE` is `"some"`, otherwise defaults to zero.
    - `STD_LFC` (logical scalar): whether log-fold changes should be standardized by the variance.
      Only possible when t-test is being used, and it is equivalent to Cohen's $d$.
      Standardized log-fold changes may be more appealing for visualization as it avoids large fold changes due to
      large variance. The choice of `STD_LFC` does not affect the calculation of the p-values.
      Note: for compatibility reasons, `STD_LFC` parameter is set, but not used in `PARAMS_WILCOX` and `PARAMS_BINOM`.
  - Reporting parameters:
    - `TOP_N_HEATMAP` (positive integer scalar or `.inf`): for how many top N markers to generate heatmaps.
      The column used for determination of the top N markers is set by `TOP_N_WT_HEATMAP`.
      You can also use all markers, either by supplying a huge number, or by using `.inf`, which is translated to R's `Inf`.
    - `TOP_N_WT_HEATMAP` (character scalar: `"top"` | `"fdr"` | `"lfc"` | `"auc"`): which column from a marker table to
      use for determination of the top N markers.
      - `"top"` can be only used when `PVAL_TYPE` is `"some"`.
      - `"auc"` can be only used inside `PARAMS_WILCOX`.
    - `TOP_N_PLOT` (positive integer scalar): same as `TOP_N_HEATMAP`, but used for marker plots.
    - `TOP_N_WT_PLOT` (character scalar: `"top"` | `"fdr"` | `"lfc"` | `"auc"`):
      same as `TOP_N_WT_HEATMAP`, but used for marker plots.

***

```yaml
CLUSTER_MARKERS_SOURCES:
  - markers_cluster_graph_leiden_r0.4:
      source_column: "cluster_graph_leiden_r0.4"
      description: "Cluster markers for Leiden clustering (r = 0.4)"
    markers_cluster_graph_leiden_r0.8:
      source_column: "cluster_graph_leiden_r0.8"
      description: "Cluster markers for Leiden clustering (r = 0.8)"
```

**Type:** list of named lists

This parameter is used to specify cell groupings ("sources") in which markers are searched for, and test and output parameters.
You can use any categorical column - results of cell clustering, `phase` (cell cycle), etc., or custom groupings defined
by `CELL_GROUPINGS` in `02_norm_clustering.yaml` and `02_int_clustering.yaml`.
See [clustering targets](#clustering) for a list of clustering names.

Let's examine the first entry (`markers_cluster_graph_leiden_r0.4`):

- `markers_cluster_graph_leiden_r0.4`: name of this comparison. You can use whatever name you want.
- `source_column: "cluster_graph_leiden_r0.4"` (character scalar):
  which cell grouping to use for computation of cell markers.
- `description: "Cluster markers for Leiden clustering (r = 0.4)"` (character scalar):
  description of this comparison, will appear in reports and plots.

In each entry you can override parameters in `CLUSTER_MARKERS_SOURCES_DEFAULTS` (by using lowercase names).
For example, let's consider the following scenario:

```yaml
CLUSTER_MARKERS_SOURCES:
  - markers_cluster_graph_leiden_r0.4:
      source_column: "cluster_graph_leiden_r0.4"
      description: "Cluster markers for Leiden clustering (r = 0.4)"
      common_params:
        plot_dimreds: ["umap", "pca"]
      params_t:
        pval_type: "some"
        top_n_wt_heatmap: "fdr"
        top_n_wt_plot: "fdr"
```

Here, `train_params` and `common_params` will override corresponding parameters in `PARAMS_T` and `COMMON_PARAMS`,
respectively, in `CELL_ANNOTATION_SOURCES_DEFAULTS`.

> Watch out for the proper indentation. See the "Merging of nested named lists" section in `vignette("scdrake_config")`.

***

```yaml
MAKE_CLUSTER_MARKERS_PLOTS: True
```

**Type:** logical scalar

Set to `False` to skip making of marker plots. In that case, for compatibility reasons, empty files will be created.

***

#### Input files

```yaml
CLUSTER_MARKERS_TABLE_TEMPLATE_RMD_FILE: "Rmd/common/cluster_markers_table_template.Rmd"
CLUSTER_MARKERS_REPORT_RMD_FILE: "Rmd/common/cluster_markers.Rmd"
```

**Type:** character scalar

Paths to RMarkdown files used for HTML reports of this pipeline stage.

***

#### Output files

```yaml
CLUSTER_MARKERS_BASE_OUT_DIR: "cluster_markers"
```

**Type:** character scalar

A path to base output directory for this stage. It will be created under `BASE_OUT_DIR` specified in `00_main.yaml` config.

***

```yaml
CLUSTER_MARKERS_REPORT_HTML_FILE: "cluster_markers.html"
CLUSTER_MARKERS_HEATMAPS_OUT_DIR: "cluster_markers_heatmaps"
CLUSTER_MARKERS_PLOTS_BASE_OUT_DIR: "cluster_markers_plots"
CLUSTER_MARKERS_DIMRED_PLOTS_BASE_OUT_DIR: "cluster_markers_dimred_plots"
CLUSTER_MARKERS_TABLES_OUT_DIR: "cluster_markers_tables"
```

**Type:** character scalar

Names of files and directories created under `CLUSTER_MARKERS_BASE_OUT_DIR`. Subdirectories are not allowed.

### Outputs

Here you can find description of the most important targets for this stage.
However, for a full overview, you have to inspect the
[source code](https://github.com/bioinfocz/scdrake/blob/main/R/plans_common.R) of the
`get_cluster_markers_subplan()` function.

#### `SingleCellExperiment` objects

`sce_dimred_cluster_markers`, `sce_final_cluster_markers`:
SCE objects with computed dimensionality reductions. Used for generation of marker plots and heatmaps

`sce_cluster_markers`: Final SCE object with cell clusterings from the `02_norm_clustering`
(single-sample pipeline) or `02_int_clustering` stage (integration pipeline). Used to compute cluster markers.
You can inspect the cell groupings which can be used for marker computation:

```{r, eval = FALSE}
drake::loadd(sce_final_cluster_markers)
SingleCellExperiment::colData(sce_final_cluster_markers)
```

#### Tibbles with parameters

Usually, each row of these tibbles is passed to a function within the `{drake}` plan.

`cluster_markers_params`: all parameters for computation of cluster markers

The following tibbles are derived from `cluster_markers_params`:
`cluster_markers_test_params`, `cluster_markers_heatmap_params`, `cluster_markers_plot_params`,
`cluster_markers_dimred_plot_params`

#### Tibbles with cluster markers test results

`cluster_markers_raw`: this tibble has the same columns as `cluster_markers_test_params`, except the list column
(`<DataFrame>`) `markers` with test results is added, and each level of `source_column` is expanded to a separate row.

So for example, if a row has `group_level = 1`, its `markers` `DataFrame` contains test results of this level versus
all other ones.

See `scran::combineMarkers()` for more details on the output format, and `scran_markers()` for a description of
changes made to the output of the former function.

***

`cluster_markers`: same as `cluster_markers_raw`, but for Wilcox tests, combined LFC is added to `DataFrame`s in
`markers` column

***

`cluster_markers_processed`: same as `cluster_markers`, but `markers` `DataFrame`s don't contain nested `DataFrame`s
(`lfc_*` or `auc_*`) - those are replaced by combined effect sizes. This way you can inspect the effect sizes of
comparisons with all other levels.

`cluster_markers_out`: holds the same data as `cluster_markers_processed`, but `markers` are coerced to dataframes
and column names are normalized to snake_case

***

`cluster_markers_for_tables`: a summarized tibble glued from `cluster_markers_out`, `cluster_markers_heatmaps_df`, and
`cluster_markers_plots_top`. Marker tables in `markers` column are in publish-ready forms for HTML output,
e.g. there are `<a>` links to ENSEMBL or PDF files of marker plots.

#### Heatmaps

`seu_for_cluster_markers_heatmaps`: a `Seurat` object used to generate heatmaps through `marker_heatmap()`
(a wrapper around `Seurat::DoHeatmap()`). In `scale.data` slot of the `RNA` assay are UMI counts transformed to z-score.

***

`cluster_markers_heatmaps_df`: a tibble derived from `cluster_markers_heatmaps_df`, but enriched with cluster markers
test results

Two heatmaps for each row in `cluster_markers_heatmaps_df` are made: one with log2 UMI counts, and second with the
same values transformed to z-score

For storage reasons, heatmap objects are not saved as R objects, but only exported to PDF.
However, you can obtain a tibble holding these objects with

```{r, eval = FALSE}
drake::loadd(cluster_markers_heatmaps_df)
heatmaps_tbl <- marker_heatmaps_wrapper(
  seu = seu_for_cluster_markers_heatmaps,
  params = cluster_markers_heatmaps_df,
  marker_type = "global",
  save = FALSE,
  return_type = "tibble"
)
```

In `heatmaps_tbl`, two list columns are appended to `cluster_markers_heatmaps_df`:

- `heatmaps`: a named list with heatmap object (`p_heatmap`) and its z-score-transformed version (`p_heatmap_zscore`).
- `markers_top`: a dataframe with top markers used in the heatmaps (subsetted `cluster_markers_processed$markers`).

For more control you can use the `marker_heatmap()` function (of which is `marker_heatmaps_wrapper()` a wrapper).

#### Marker plots

`cluster_markers_plots_top`: a tibble with markers for which plots will be made.
See `?markers_plots_top` for details.

As for heatmaps, plot objects (`{ggplot2}`, `{patchwork}`) are not saved, but the underlying function can be used:

```{r, eval = FALSE}
markers_plots_files(sce_dimred_cluster_markers, cluster_markers_plots_top, save = FALSE, return_type = "tibble")
```

Similarly, the actual plotting function can be used, see `?marker_plot`

#### Dimensionality reduction plots

`cluster_markers_dimred_plots`: a tibble with dimensionality reduction plots for each of `source_column` in
the `CLUSTER_MARKERS_SOURCES` parameter and `PLOT_DIMREDS` in `COMMON_PARAMS`

#### Other targets

`config_cluster_markers`: a list holding parameters for this stage
