---
title: "Input and quality control stage (`01_input_qc`)"
description: "01_input_qc stage of the single-sample pipeline."
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Input and quality control stage (`01_input_qc`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

***

## {.tabset}

### Overview

An initial stage to read in scRNA-seq data, remove empty droplets (optionally), and perform quality control,
and cell and gene filtering.

More information about scRNA-seq quality control can be found in
[OSCA](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html).

`r emoji::emoji("gear")` Config file: `config/single_sample/01_input_qc.yaml`

`r emoji::emoji("clipboard")` HTML report target (in `config/pipeline.yaml`): `DRAKE_TARGETS: ["report_input_qc"]`

[`r emoji::emoji("scroll")` Example report](https://onco.img.cas.cz/novotnyj/scdrake/single_sample/pbmc1k/01_input_qc/01_input_qc.html)
([used config](https://github.com/bioinfocz/scdrake/blob/main/tests/testthat/run_pipeline_config_patches/single_sample_pbmc1k_full/01_input_qc.default.yaml))

`r emoji::emoji("ladder")` Structure

- Reading in feature-barcode matrix either from:
  - `cellranger` output
  - `SingleCellExperiment` object as a Rds file or as a target from an existing `{drake}` cache (i.e. from a different
    `{scdrake}` project)
  - Delimited table
- Subsetting of the imported `SingleCellExperiment` object to cells of interest based on their metadata values,
  e.g. a set of clusters from a particular clustering method (optional)
- Removal of [empty droplets](https://bioconductor.org/books/3.15/OSCA.advanced/droplet-processing.html)
  (`DropletUtils::emptyDrops()`) (optional)
- Calculation of per-cell QC metrics (`scater::perCellQCMetrics()`,
  [theory](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html#common-choices-of-qc-metrics)):
  - Number of UMI
  - Number of detected genes (non-zero UMI count)
  - Percentage of expressed mitochondrial genes ($\frac {UMI_{mitochondrial}} {UMI_{sum}} * 100$)
- Two methods for cell filtering (optional):
  - Dataset-sensitive filtering, based on the median absolute deviation (MAD) from the median value of
     each QC metric across all cells
     ([theory](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html#quality-control-outlier)).
     For "number of UMI" metric, only a lower tail is used.
  - Custom filtering, based on fixed thresholds. For "number of UMI" metric, both upper and lower bounds are used
    ([theory](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html#fixed-qc).
  - For both filtering types you can choose how to join the filters: either jointly (using the *AND* operator) or
    individually (using the *OR* operator).
- Gene filtering (optional) based on a minimum number of UMI per cell and a minimum ratio of cells expressing a gene.
  - A gene is considered expressed when number of its UMI across all cells is greater than X and
    at the same time it is expressed in at least Y ratio of cells.
- Plots of QC metrics with visualization of differences between dataset-sensitive and custom filtering
- Preparation of gene annotation
- Final selection of either dataset-sensitive or custom filtered dataset. This selection is upon the user.

### Config parameters

Config for this stage is stored in `config/single_sample/01_input_qc.yaml` file.
Directory with this file is read from the `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` environment variable upon `{scdrake}` load or
attach, and saved as `scdrake_single_sample_config_dir` option. This option is used as the default argument value in
several `{scdrake}` functions.

***

#### Input files

```yaml
INPUT_DATA:
  type: "cellranger"
  path: "/path/to/dir"
  delimiter: ","
  target_name: "target_name"
```

**Type:** list of named character scalars

This parameter is specifying input data type and path. There are four possible types of input data:

- `type: "cellranger"`: an output from
  [cellranger](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger),
  that is, files of a raw feature-barcode matrix - `barcodes.tsv`, `features.tsv`, and `matrix.mtx`
  (can be also gzipped having a `.gz` extension). Internally, BC matrix is imported via `DropletUtils::read10xCounts()`
  - You can download example PBMC data from 10x Genomics using the `download_pbmc1k()` and `download_pbmc1k()` functions.
- `type: "table"`: a delimited file (table) representing a feature-barcode matrix, with an additional column named
  `ENSEMBL` containing Ensembl IDs of features (rows).
- `type: "sce"`: a saved `SingleCellExperiment` in Rds format (i.e. saved by the `saveRDS()` function).
  The object must contain feature-barcode matrix with raw UMI counts in the `counts` slot of `assays`
  (such that this BC matrix can be retrieved by `counts(sce)`). 
  - Additionaly, you can include whatever you want in `colData()` and use it later (e.g. cell grouping/clusters
    to compute markers or to include batch effect during that).
- `type: "sce_drake_cache"`: a saved `SingleCellExperiment` in a `drake` cache, e.g. from an another `scdrake` project

- `path: "path/to/file/or/dir"`: a path to input file or directory (can be relative to project root directory).
  - For `type: "cellranger"`, this is a path to directory containing the `barcodes.tsv`, `features.tsv`, and `matrix.mtx`
  files (can be also gzipped having a `.gz` extension)
  - For `type: "sce_drake_cache"`, a path to `drake` cache directory (usually named as `".drake"`)
  - For `type: "sce"` or `"table"`, a path to Rds or text file

- `delimiter: ","`: used when `type: "table"`. Specifies the field delimiter in the table.

- `target_name: "target_name"`: used when `type: "sce_drake_cache"`. Specifies a name of target in `drake` cache to be
  imported, e.g. `"sce_final_input_qc"`

***

```yaml
INPUT_QC_REPORT_RMD_FILE: "Rmd/single_sample/01_input_qc.Rmd"
```

**Type:** character scalar

A path to RMarkdown file used for HTML report of this pipeline stage. For spatial extention, the default RMarkdown file is `01_input_qc_spatial.Rmd`

***

#### Subsetting of imported data

```yaml
INPUT_DATA_SUBSET: null
## Example of simple subsetting by cluster numbers
INPUT_DATA_SUBSET:
  subset_by: "cluster_kmeans_k6"
  values: ["3", "4"]
  negate: false
```

**Type:** `null` (default) or named list

The imported data can be subsetted by a simple selection of values present in a column of cell metadata (`colData()`).
In the example above, cells assigned to clusters `"3"` and `"4"` of the `cluster_kmeans_k6` columns will be kept.
You can also negate the selection by specifying `negate: true`.

***

#### Spatial extention


```yaml
SPATIAL: False
```

**Type:** logical scalar

If `True`, pipeline enables spatial extension.

In the input_qc stage, the spatial extension consists only in enabling pseudo tissue visualization, and in adding coordinates (array_col and array_row from tissue_positions.csv) to the Cell Metadata. 

***

```yaml
SPATIAL_LOCKS: Null
```

**Type:** Null or character scalar

A path to tissue_position.csv from SpaceRanger spatial output folder used for adding coordinates (array_col and array_row) to Cell Metadata. Only void when SPATIAL is enabled.

***

#### Removal of empty droplets

See `?DropletUtils::emptyDrops` for more details.

***

```yaml
EMPTY_DROPLETS_ENABLED: True
```

**Type:** logical scalar

If `False`, skip calculation and removal of empty droplets.

You might consider turning off this procedure if the input is not a raw feature-barcode matrix from `cellranger`,
but an already processed dataset (table or `SingleCellExperiment` object).
Otherwise `DropletUtils::emptyDrops()` will fail as there are not enough empty droplets having the total UMI count
< `EMPTY_DROPLETS_LOWER` (100 by default).

***

```yaml
EMPTY_DROPLETS_LOWER: 100
```

**Type:** positive integer scalar

A lower bound on the total UMI count at or below which all barcodes are assumed to correspond to empty droplets.

***

```yaml
EMPTY_DROPLETS_FDR_THRESHOLD: 0.01
```

**Type:** numeric scalar <0; 1>

A threshold for FDR adjusted p-values for null hypothesis that barcode comes from ambient environment.
In other words, probability that a droplet was empty and contained ambient RNA.

#### Cell filtering

Two types of cell filtering are performed: *dataset-sensitive* (using MAD threshold),
and *custom* (using custom thresholds).

In both filtering types, violation of only one metric threshold leads to removal of a cell.

The choice of cell filtering thresholds depends very much on your dataset. What works for an ordinary high-quality dataset
might not work well for a low-quality or special dataset with rare cell types.
The [quality control](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html) chapter in OSCA gives
an excellent background to the QC of scRNA-seq data.

***

```yaml
ENABLE_CELL_FILTERING: True
```

**Type:** logical scalar

If `False`, both types of cell filtering will be disabled.

The filtering thresholds in both filtering types will be overridden such that every cell will pass the filtering
(e.g. `MAD_THRESHOLD: .inf` or `MIN_UMI_CF: -.inf`). In the end, `SAVE_DATASET_SENSITIVE_FILTERING` will have no
effect as both filtered SCE targets will be identical. All QC metrics and plots will still be calculated and created.

***

```yaml
SAVE_DATASET_SENSITIVE_FILTERING: True
```

**Type:** logical scalar

If `True`, proceed to other stages with dataset filtered by dataset-sensitive filtering (target `sce_qc_filter_genes`),
otherwise with dataset filtered by custom filtering (target `sce_custom_filter_genes`).
Note that the selected SCE target will be referred to as `sce_final_input_qc` in subsequent stages.

##### Dataset-sensitive cell filtering

```yaml
MAD_THRESHOLD: 3
```

**Type:** positive numeric scalar

A threshold for maximum MAD (median absolute deviation) for QC cell metrics:

- Number of UMI (lower tail).
- Number of detected genes = non-zero UMI count (lower tail).
- Percentage of expressed mitochondrial genes = $\frac {UMI_{mitochondrial}} {UMI_{sum}} * 100$ (upper tail).

MAD threshold of 3 will retain 99% of non-outlier values that follow a normal distribution.

"Lower tail" means cells having a metric value less than `-MAD_THRESHOLD` MAD will be discarded.
For "upper tail", it is `+MAD_THRESHOLD` MAD.

Violation of only one metric threshold leads to removal of a cell.

To disable the dataset-sensitive filtering, set `MAD_THRESHOLD: .inf`.
That will force passing of each cell as every QC metric will be always lower than positive infinity MAD.

***

```yaml
DATASET_SENSITIVE_FILTERS_OPERATOR: "&"
```

**Type:** a character scalar (`"&" | "|"`)

How to join the QC filters:

- Jointly (*AND*/`&` operator), i.e., remove only cells that violate **ALL** filters (permissive)
- Individually (*OR*/`|` operator), i.e., remove cells that violate **AT LEAST ONE** filter (strict)

##### Custom cell filtering

```yaml
MIN_UMI_CF: 1000
```

**Type:** positive integer scalar

A threshold for minimum number of UMI per cell, i.e. cells with UMI less than `MIN_UMI_CF` will be removed.

To disable this filter, set `MIN_UMI_CF: -.inf`

***

```yaml
MAX_UMI_CF: 50000
```

**Type:** positive integer scalar

A threshold for maximum number of UMI per cell, i.e. cells with UMI greater than `MAX_UMI_CF` will be removed.

To disable this filter, set `MAX_UMI_CF: .inf`

***

```yaml
MIN_FEATURES: 1000
```

**Type:** positive integer scalar

A threshold for minimum number of features (genes) detected per cell,
i.e. cells with detected features less than `MIN_FEATURES` will be removed.

To disable this filter, set `MIN_FEATURES: -.inf`

***

```yaml
MAX_MITO_RATIO: 0.2
```

**Type:** numeric scalar <0; 1>

A threshold for maximum ratio of expressed mitochondrial genes per cell,
i.e. cells with mitochondrial genes detected in more than (`MAX_MITO_RATIO` * 100)% of all genes will be removed.

To disable this filter, set `MAX_MITO_RATIO: .inf`

***

```yaml
CUSTOM_FILTERS_OPERATOR: "&"
```

**Type:** a character scalar (`"&" | "|"`)

How to join the QC filters:

- Jointly (*AND*/`&` operator), i.e., remove only cells that violate **ALL** filters (permissive)
- Individually (*OR*/`|` operator), i.e., remove cells that violate **AT LEAST ONE** filter (strict)

#### Gene filtering

Gene filtering thresholds are applied in both types of cell filtering (after it is performed).
The filter is computed by the `get_gene_filter()` function as following:

```{r, eval = FALSE}
num_cells <- min_ratio_cells * ncol(sce)
is_expressed <- rowSums(counts(sce) >= min_umi) >= num_cells
```

A gene is considered expressed when number of its UMIs across all cells is greater than `min_umi` and
at the same time it is expressed in at least `min_ratio_cells` ratio of cells.

***

```yaml
ENABLE_GENE_FILTERING: True
```

**Type:** logical scalar

If `False`, gene filtering will be disabled.

The filtering thresholds will be overridden such that every gene will pass the filtering
(`MIN_UMI: 0` and `MIN_RATIO_CELLS: 0`).

***

```yaml
MITO_REGEX: "^MT-"
RIBO_REGEX: "^RP[SL]"
```

**Type:** character scalar

Regexes used to match and count occurences of mitochondrial and ribosomal features.
The latter is currently not used for gene filtering.

***

```yaml
MIN_UMI: 1
```

**Type:** zero or positive integer scalar

A threshold for minimum number of UMI per cell, i.e. genes with UMI < `MIN_UMI` will be removed.

***

```yaml
MIN_RATIO_CELLS: 0.01
```

**Type:** numeric scalar <0; 1>

A minimum ratio of cells expressing a gene,
i.e. genes expressed in less than (`MIN_RATIO_CELLS` * 100)% of cells will be removed.

***

#### Output files

```yaml
INPUT_QC_BASE_OUT_DIR: "01_input_qc"
```

**Type:** character scalar

A path to base output directory for this stage. It will be created under `BASE_OUT_DIR` specified in `00_main.yaml` config.

***

```yaml
INPUT_QC_REPORT_HTML_FILE: "01_input_qc.html"
```

**Type:** character scalar

A name of HTML report file for this stage. Created in `INPUT_QC_BASE_OUT_DIR`. Subdirectories are not allowed.

#### HTML output parameters

```yaml
INPUT_QC_KNITR_MESSAGE: False
INPUT_QC_KNITR_WARNING: False
INPUT_QC_KNITR_ECHO: False
```

**Type:** logical scalar

These are passed to `knitr::opts_chunk()` and used for rendering of stage's HTML report.

### Outputs

Here you can find description of the most important targets for this stage.
However, for a full overview, you have to inspect the
[source code](https://github.com/bioinfocz/scdrake/blob/main/R/plans_single_sample.R) of the
`get_input_qc_subplan()` function.

Targets can be loaded from the `{drake}` cache (`.drake` directory by default) using the `drake::loadd()` or
`drake::readd()` functions.

HTML report target name: `report_input_qc`

#### `SingleCellExperiment` objects

`sce_raw`: an untouched SCE object as loaded from:

- `DropletUtils::read10xCounts()` (`cellranger` input)
- `readr::read_delim()` (delimited table input)
- `SingleCellExperiment` object (Rds file or `{drake}` cache)

***

`sce_valid_cells`: a SCE object with empty droplets removed

***

`sce_unfiltered`: `sce_valid_cells` with added columns:

- Columns from the `cell_qc` target containing DataFrame with cell QC metrics
- `discard_qc`: dataset-sensitive filter
- `discard_custom`: custom filter

***

`sce_qc_filter`, `sce_qc_filter_genes`: a SCE object with cells filtered by dataset-sensitive filtering.
The latter with filtered genes.

`sce_custom_filter`, `sce_custom_filter_genes`: a SCE object with cells filtered by custom filtering.
The latter with filtered genes.

***

`sce_selected`: a SCE object selected from either
`sce_qc_filter_genes` or `sce_custom_filter_genes` - depends on the `SAVE_DATASET_SENSITIVE_FILTERING` parameter

`sce_final_input_qc`: `sce_selected` with added gene annotation (`gene_annotation` target) to `rowData()`.
This is the final SCE object which will be used in the next stage `norm_clustering`.

#### Cell filters

`cell_qc`: a DataFrame with cell QC metrics

***

The following list of filters are computed by `scater::isOutlier()` from `cell_qc` columns,
which are shown for individual filters (e.g. `qc_lib`: `total` column).
In those, `FALSE` refers to cells **not passing** a filter.

`qc_filters`: a list of named lists with logical values for dataset-sensitive cell filtering:

- `qc_lib` (`total`): low number of UMI (lower tail)
- `qc_nexprs` (`detected`): low number of detected genes = non-zero UMI count (lower tail)
- `qc_mito` (`subsets_mito_percent`): high percentage of expressed mitochondrial genes expression (upper tail)

`custom_filters`: a list of named lists with logical values for custom cell filtering:

- `low_count` (`total`): low number of UMIs.
- `high_count` (`total`): high number of UMIs.
- `low_expression` (`detected`): low number of detected genes.
- `high_mito` (`subsets_mito_percent`): high percentage of expressed mitochondrial genes.

`qc_filter`, `custom_filter`: a summarized filter (by "or" operation) of `qc_filters` and `custom_filters`, respectively

#### Gene filters

`sce_qc_gene_filter`, `sce_custom_gene_filter`: gene filters (logical vectors) obtained from `get_gene_filter()` for
SCE objects with cells filtered by dataset-sensitive and custom filter, respectively.
In those, `FALSE` refers to genes **not passing** the filter.

#### Plots

`sce_unfiltered_plotlist`: a list of various violin plots for cell QC metrics (`{ggplot2}` objects), generated using
`sce_unfiltered` target. Cells are colored by `discard_qc` column - blue points pass the dataset-sensitive filter,
while the orange ones do not.

For example, the following code is used to plot total UMI counts:

```{r, eval = FALSE}
scdrake::plot_colData(
  sce_unfiltered,
  y = "total",
  colour_by = "discard_qc",
  title = "Total count",
  scale_y = ggplot2::scale_y_log10()
)
```

***

`sce_qc_filter_genes_plotlist`, `sce_custom_filter_genes_plotlist`: similar to `sce_unfiltered_plotlist`,
but in each of them, cells are colored by the other filter. That is, cells in `sce_qc_filter_genes_plotlist` are
colored by `discard_custom` column, and thus, showing which cells would be discarded in custom-filtered dataset.
Vice versa for `sce_custom_filter_genes_plotlist`.

#### Other targets

`config_input_qc`: a list holding parameters for this stage

***

`barcode_ranks`: a DataFrame with barcode ranks for
[knee plot](https://bioconductor.org/books/3.15/OSCA.advanced/droplet-processing.html),
computed by `DropletUtils::barcodeRanks()` on UMI counts from `sce_raw` target

`empty_droplets`: a DataFrame of statistics for empty droplets, computed `DropletUtils::emptyDrops()` on UMI counts
from `sce_raw` target

***

`sce_history`: a tibble with cell and gene "history", showing their numbers in unfiltered and filtered data.

`sce_history_plot`: a `{ggplot2}` object summarizing information from `sce_history` target.

***

`gene_annotation`: a dataframe with gene annotation, processed such that:

- If a single ENSEMBL ID has multiple symbols, gene descriptions, or ENTREZ IDs, they are collapsed by comma (`,`).
- ENSEMBL ID is used as a symbol for ENSEMBL IDs with unknown symbols.
- ENSEMBL ID is appended to symbols having multiple ENSEMBL IDs.
  - For example, *TBCE* has both ENSG00000285053 and ENSG00000284770 ENSEMBL IDs assigned ->
    its symbol is changed to TBCE_ENSG00000285053 and TBCE_ENSG00000284770.

`gene_annotation` can be also retrieved from `rowData(sce_final_input_qc)`
