---
title: "Normalization and clustering stage"
description: "02_norm_clustering stage of the single-sample pipeline."
date: "`r Sys.Date()`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Normalization and clustering stage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

***

For a general overview of this single-sample pipeline stage see `vignette("pipeline_overview")`.

Config for this stage is stored in `02_norm_clustering.yaml` file.
Directory with this file is read from `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` environment variable upon `{scdrake}` load or
attach, and saved as `scdrake_single_sample_config_dir` option. This option is used as the default argument value in
several `{scdrake}` functions.

***

# Config parameters

## Normalization parameters

`NORMALIZATION_TYPE: "scran"`

**Type:** character scalar (`"scran"` | `"sctransform"` | `"none"`)

A normalization type:

- `"scran"`: normalization by deconvolution.
- `"sctransform"`: regularized negative binomial regression.
- `"none"`: skip normalization. This should be used only in case you are importing a `SingleCellExperiment`
  object in the `01_input_qc` stage and this object was normalized before within this pipeline,
  i.e. you are importing a modified `sce_norm` target or one of its downstream, dependent `SingleCellExperiment` targets.
  - For example, you can take `sce_final_norm_clustering`, subset cells to e.g. three clusters, save the result to Rds
    file, and run the pipeline again while importing this file in the `01_input_qc` stage.
    Note that in that case you should disable empty droplets removal, and cell and gene filtering,
    as data are already of high quality. Overall, this is an example of reclustering, which is otherwise hard to
    achieve within the pipeline.

See `vignette("pipeline_overview")` for more details about normalization methods.

### `scran` normalization parameters

These parameters are used when `NORMALIZATION_TYPE` is `"scran"`.

***

`SCRAN_USE_QUICKCLUSTER: True`

**Type:** logical scalar

Whether to cluster cells prior to normalization. See `?scran::quickCluster` for more details.

***

`SCRAN_QUICKCLUSTER_METHOD: "igraph"`

**Type:** character scalar (`"igraph"` | `"hclust"`)

A clustering method: `"igraph"` uses graph-based clustering, while `"hclust"` uses hierarchical clustering.
See `?scran::quickCluster` for more details.

### `sctransform` normalization parameters

These parameters are used when `NORMALIZATION_TYPE` is `"sctransform"`.

***

`SCT_VARS_TO_REGRESS: null`

**Type:** list of character scalar or `null`

Batch variables to regress out during the `SCTransform` normalization. You can regress out variables such as
`["percent.mt", "nFeature_RNA", "nCount_RNA"]` or cell cycle using `["S.Score", "G2M.Score"]` or `["CC.Difference"]`.

However, for correction of cell cycle effect, a better method is the removal of cell cycle-related genes during
selection of highly variable genes (HVGs).
This can be set by [HVG selection parameters](#highly-variable-genes-hvgs-selection), and the approach is explained
in more details in `vignette("pipeline_overview")`.

***

`SCT_N_HVG: 3000`

**Type:** integer scalar

A number of top HVGs to return from the `SCTransform` normalization.

## Highly variable genes (HVGs) selection

`HVG_METRIC: "gene_var"`

**Type:** character scalar (`"gene_var"` | `"gene_cv2"` | `"sctransform"`)

A metric used to find HVGs. See <https://bioconductor.org/books/release/OSCA.basic/feature-selection.html> for more details.

- `"gene_var"`: variance of the log-counts (`scran::modelGeneVar()`).
- `"gene_cv2"`: coefficient of variation (`scran::modelGeneCV2()`).
- `"sctransform"`: use HVGs selected by `SCTransform`.
  Only relevant when the `NORMALIZATION_TYPE` parameter is `"sctransform"`.
  You can adjust number of these HVGs by the `SCT_N_HVG` parameter.

***

`HVG_SELECTION: "top"`

**Type:** character scalar (`"top"` | `"significance"` | `"threshold"`)

A HVG selection strategy. Only relevant when the `HVG_METRIC` parameter is `"gene_var"` or `"gene_cv2"`.

- `"top"`: take top X genes according to the metric
  ([details](https://bioconductor.org/books/release/OSCA.basic/feature-selection.html#hvg-selection)).
- `"significance"`: use a FDR threshold
  ([details](https://bioconductor.org/books/release/OSCA.advanced/more-hvgs.html#more-hvg-selection-strategies)).
- `"threshold"`: use a threshold on minimum value of the metric
  ([details](https://bioconductor.org/books/release/OSCA.advanced/more-hvgs.html#more-hvg-selection-strategies)).

For `"top"` and `"threshold"`, `bio` and `ratio` columns are used for `HVG_METRIC` of `"gene_var"` and `"gene_cv2"`,
respectively. These columns are present in the DataFrame returned from the underlying method
(`scran::modelGeneVar()` or `scran::modelGeneCV2()`).

***

`HVG_SELECTION_VALUE: 1000`

**Type:** depends on value of the `HVG_SELECTION` parameter:

- A positive integer scalar for `HVG_SELECTION: "top"`.
- A numeric scalar $<0; 1>$ for `HVG_SELECTION: "significance"`.
- A numeric scalar for `HVG_SELECTION: "threshold"`.

A threshold value for HVG selection. The following parameter combinations are recommended defaults:

- Top 1000 genes by a HVG metric.

```yaml
HVG_SELECTION: "top"
HVG_SELECTION_VALUE: 1000
```

- Genes with a HVG metric significance (FDR) lesser than `0.05`.

```yaml
HVG_SELECTION: "significance"
HVG_SELECTION_VALUE: 0.05
```

- Genes with a HVG metric value greater than `0` or `1`.

```yaml
HVG_METRIC: "gene_var"
HVG_SELECTION: "threshold"
HVG_SELECTION_VALUE: 0
```

```yaml
HVG_METRIC: "gene_cv2"
HVG_SELECTION: "threshold"
HVG_SELECTION_VALUE: 1
```

### Removal of cell cycle-related genes prior to HVG selection

`HVG_RM_CC_GENES: False`

**Type:** logical scalar

Whether to apply a correction for cell cycle prior to HVG selection.
We implement a cell cycle-related genes removal strategy described
[here]https://bioconductor.org/books/release/OSCA.advanced/cell-cycle-assignment.html#removing-cell-cycle-related-genes).
It is based on the percentage of variance explained by the cell cycle phase in the expression profile for each gene.

Note that when `SCTransform` is used for normalization including cell cycle variables to regress out, than
the results of this strategy remain unclear.

***

`HVG_CC_GENES_VAR_EXPL_THRESHOLD: 5`

**Type:** positive numeric scalar

A threshold on gene cell cycle variance explained, i.e. genes with var. expl. greater than this threshold will be
marked as cell cycle-related and removed *prior* to selection of HVGs.

## Doublet detection and filtering

Doublets are droplets in which multiple cells were captured.

***

`MAX_DOUBLET_SCORE: 3.5`

**Type:** positive numeric scalar or `null`

A threshold for doublet score computed by `scDblFinder::computeDoubletDensity()`.
Cells with doublet score greater than the threshold will be removed.
If `null`, doublets are not removed, but doublet score is still computed.

## Dimensionality reduction

### PCA

`PCA_SELECTION_METHOD: "forced"`

**Type:** character scalar (`"forced"` | `"elbow"` | `"technical_noise"`)

A method to select first N principal components (PCs):

- `"forced"`: use the number of PCs specified by the `PCA_FORCED_PCS` parameter.
- `"elbow"`: use the
  [elbow point method](https://bioconductor.org/books/release/OSCA.advanced/dimensionality-reduction-redux.html#using-the-elbow-point).
- `"technical_noise"`: use the
  [technical noise](https://bioconductor.org/books/release/OSCA.advanced/dimensionality-reduction-redux.html#using-the-technical-noise).

***

`PCA_FORCED_PCS: 15`

**Type:** positive integer scalar

If the `PCA_SELECTION_METHOD` parameter is `"forced"`, use this number of first principal components.

## t-SNE

[t-stochastic neighbor embedding](https://bioconductor.org/books/release/OSCA.basic/dimensionality-reduction.html#t-stochastic-neighbor-embedding)

See `?scater::runTSNE` for the underlying method.

***

`TSNE_PERP: 20`

**Type:** numeric scalar

t-SNE perplexity.

***

`TSNE_MAX_ITER: 1000`

**Type:** positive integer scalar

A maximum number of t-SNE iterations.

## Clustering

### Graph-based clustering

Parameters for [graph-based clustering](https://bioconductor.org/books/release/OSCA.basic/clustering.html#clustering-graph).

***

`GRAPH_SNN_N: 10`

**Type:** positive integer scalar

A number of shared nearest neighbours. Passed as `k` parameter to `bluster::makeSNNGraph()`.

***

`GRAPH_SNN_TYPE: "rank"`

**Type:** character scalar (`"rank"` | `"number"` | `"jaccard"`)

A type of weighting scheme to use for shared neighbors. Passed as `type` parameter to `bluster::makeSNNGraph()`.

### k-means clustering

`KMEANS_K: !code 3:6`

**Type:** integer vector

A vector of numbers clusters for which to compute k-means.

### SC3 clustering

`SC3_K: !code 5:6`

**Type:** integer vector

A vector of numbers clusters for which to compute SC3 clustering.

***

`SC3_N_CORES: 8`

**Type:** integer scalar

A number of CPU cores to use for SC3 computation.

***

`SC3_DRY: False`

**Type:** logical scalar

If `True`, just generate random cluster assignments (for testing purposes).

## Cell grouping assignment

```yaml
CELL_GROUPINGS:
  - cluster_graph_louvain_annotated:
      source_column: "cluster_graph_louvain"
      description: "Graph-based clustering (Louvain alg.), annotated clusters"
      assignments:
        1: "memory_CD4+"
        6: "B"
        7: "memory_CD4+"
    cluster_kmeans_kc_3_renamed:
      source_column: "cluster_kmeans_kc_3"
      description: "k-means clustering (k = 3), renamed clusters"
      assignments:
        1: "kc_1"
        2: "kc_2"
    cluster_sc3_6_renamed:
      source_column: "cluster_sc3_6"
      description: "SC3 clustering (k = 6), renamed clusters"
      assignments:
        1: "sc3_1"
        5: "sc3_5"
```

**Type:** list of named lists or `null`

This parameter specifies new cell groupings based on existing columns (primarily the clustering ones). Each key is a
name of new cell grouping (column in `colData()` of the `sce_dimred` target) and contains a named list in the form:

- `source_column` (character scalar): a name of existing column in `colData()`.
- `description` (character scalar, optional): a description of the cell grouping, will appear in plots.
  If not set, the name of the assignment will be used.
- `assignments` (named list/vector of character scalars):
  - `<old level>: "<new level>"`: assignment of an old level to a new level. Different old levels with the same new
    level will be merged. Order of assignments does not matter, levels are just replaced
    (under the hood, `dplyr::recode()` is used). Unspecified levels will be kept as they are.
    You cannot replace existing groupings (columns).

For possible column names for `source_column`, see the `sce_final_norm_clustering` target in the [section](#targets) below.

Let's look at one of the example assignments:

```yaml
cluster_graph_louvain_annotated:
  source_column: "cluster_graph_louvain"
  description: "Graph-based clustering (Louvain alg.), annotated clusters"
  assignments:
    1: "memory_CD4+"
    6: "B"
    7: "memory_CD4+"
```

- `cluster_graph_louvain_annotated`: a name of new cell grouping added to `colData()`.
- `source_column: "cluster_graph_louvain"`: a name of column in `colData()` whose level will be used for the
  new assignments.
- `description: "Graph-based clustering ..."`: a description of the new assignment. If it was not set,
  the name of the assignment (`cluster_graph_louvain_annotated`) would be used instead.
- `assignments`: assignments of the old levels to new ones. That is, level (cluster) `6` will be renamed to `"B"`,
  and levels `1` and `7` will be merged to a new level `"memory_CD4+"`.
  All unspecified levels will be kept as they are.

NOTE: The example cluster renamings serve as an example for 10x PBMC 1k dataset, but initially on your dataset you
should use `null` and set renamings after you examine the pipeline results.

## Dimensionality reduction plots

`NORM_CLUSTERING_REPORT_DIMRED_NAMES: ["umap", "pca", "tsne"]`

**Type:** character vector (`"umap"` | `"pca"` | `"tsne"`)

A vector of names of dimensionality reduction methods for which plots in the report will be made.

***

`NORM_CLUSTERING_REPORT_CLUSTERING_NAMES: ["graph", "kmeans", "sc3"]`

**Type:** character vector (`"graph"` | `"kmeans"` | `"sc3"`)

A vector of names of clustering methods for which plots in each reduced dimension defined in the
`NORM_CLUSTERING_REPORT_DIMRED_NAMES` parameter will be made and display in the report.

- `"graph"`: graph-based clustering, two algorithms: Louvain and Walktrap.
- `"kmeans"`: k-means clustering, two algorithms: best K and numbers of clusters defined in the `KMEANS_K` parameter.
- `"sc3"`: SC3 clustering: numbers of cluster defined in the `SC3_K` parameter.

***

```yaml
NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER:
  - "phase": "Cell cycle phases"
    "doublet_score": "Doublet score"
    "total": "Total number of UMI"
    "detected": "Detected number of genes"
    "cluster_graph_louvain_annotated": null
```

**Type:** list of named lists with character scalars

Names of other variables in `colData()` to plot and color by, displayed in the report.

The format is `"<variable_name": "<description>"`, where description will appear in plot title.

Variables defined in the `CELL_GROUPINGS` parameter could have <description> set to `null` -
their description from `CELL_GROUPINGS` will be used.

## Input files

`SELECTED_MARKERS_FILE: !code system.file("extdata", "selected_markers.csv", package = "scdrake", mustWork = TRUE)`

**Type:** character scalar or `null`

A path to CSV file with groups of markers to make plots for. Set to `null` to skip.
The plots will be created for each of `NORM_CLUSTERING_REPORT_DIMRED_NAMES`.

The format is following: each row defines gene symbols for a group.
First column is a name of group and second column are gene symbols delimited by ":". Example:

`Naive_CD4+_T,IL7R:CCR7`

**Do not use a header!**

The default value returns a path to package-bundled CSV file, but normally you should use
`SELECTED_MARKERS_FILE: "path/to/file.csv"`.

***

```yaml
NORM_CLUSTERING_REPORT_RMD_FILE: "Rmd/single_sample/02_norm_clustering.Rmd"
NORM_CLUSTERING_REPORT_SIMPLE_RMD_FILE: "Rmd/single_sample/02_norm_clustering_simple.Rmd"
```

**Type:** character scalar

Paths to RMarkdown files used for HTML reports of this pipeline stage.

`NORM_CLUSTERING_REPORT_RMD_FILE` is technically more detailed, while `NORM_CLUSTERING_REPORT_SIMPLE_RMD_FILE`
outputs only the dimensionality reduction plots.

## Output files

`NORM_CLUSTERING_BASE_OUT_DIR: "02_norm_clustering"`

**Type:** character scalar

A path to base output directory for this stage. It will be created under `BASE_OUT_DIR` specified in `00_main.yaml` config.

***

```yaml
NORM_CLUSTERING_SELECTED_MARKERS_OUT_DIR: "selected_markers"
NORM_CLUSTERING_REPORT_HTML_FILE: "02_norm_clustering.html"
NORM_CLUSTERING_REPORT_SIMPLE_HTML_FILE: "02_norm_clustering_simple.html"
```

**Type:** character scalar

Names of files and directories created under `NORM_CLUSTERING_BASE_OUT_DIR`. Subdirectories are not allowed.

## HTML output parameters

```yaml
NORM_CLUSTERING_KNITR_MESSAGE: False
NORM_CLUSTERING_KNITR_WARNING: False
NORM_CLUSTERING_KNITR_ECHO: False
```

**Type:** logical scalar

These are passed to `knitr::opts_chunk()` and used for rendering of stage's HTML report.

# Targets

Here you can find description of the most important targets for this stage.
However, for a full overview, you have to inspect the
[source code](https://github.com/bioinfocz/scdrake/blob/main/R/plans_single_sample.R) of the
`get_input_norm_clustering_subplan()` function.

## `SingleCellExperiment` objects

`sce_cc`: a SCE object (the `sce_final_input_qc` target from the `input_qc` stage) with added cell cycle information
to `colData()`: `phase` (`"G1"`, `"G2m"` or `"S"`), `s_score`, `g2m_score`, `cc_difference` (`s_score` - `g2m_score`).

***

`sce_norm`: `sce_cc` with normalized counts in `SingleCellExperiment::logcounts()` assay and `normalization_type` item
added to `metadata()`.

***

`sce_norm_hvg`: `sce_norm` with added HVG information:

- `metadata()`: `hvg_metric`, `hvg_selection`, `hvg_selection_value`, `hvg_metric_fit`, `hvg_rm_cc_genes`, `hvg_ids`
- `rowData()`: `is_hvg` (logical)

If the parameter `HVG_RM_CC_GENES` is `True`, `sce_norm_hvg` will also contain a PCA dimensionality reduction
(named `pca_with_cc`) for whose computation all HVGs were used.

***

`sce_rm_doublets`: `sce_norm_hvg` with removed doublets and added information:

- `metadata()`: `has_filtered_doublets` (`FALSE` if `MAX_DOUBLET_SCORE` parameter is `null`), `max_doublet_score`
- `colData()`: `doublet_score` (numeric), `is_doublet` (logical)

***

`sce_pca`: `sce_rm_doublets` with calculated PCA (50 PCs). Reduced dimensions can be retrieved by
`SingleCellExperiment::reducedDim(sce_pca, "pca")`.

***

`sce_pca_selected_pcs`: `sce_pca` with selected number of PCs. Reduced dimensions in `pca` slot are subsetted to
the selected number of PCs, and `pca_all` slot contains the full matrix of 50 PCs. Information about selection of PCs
is added to `metadata()`: `pca_selection_method`, `pca_selected_pcs`.

***

`sce_dimred`: `sce_pca_selected_pcs` with calculated t-SNE and UMAP,
both using the selected number of PCs for calculation.

***

`sce_final_norm_clustering`: a final SCE object of this stage which will be used in the `cluster_markers` and
`contrasts` stages. This object is derived from `sce_dimred`, but cell clusterings along with `CELL_GROUPINGS` are
added to `colData()`.

The names of columns containing cell clusterings are:

- Graph-based clustering: `cluster_graph_walktrap`, `cluster_graph_louvain`
- k-means clustering: `cluster_kmeans_kbest`, `cluster_kmeans_kc_<k>` (all Ks from the `KMEANS_K` parameter)
- SC3 clustering: `cluster_sc3_<k>` (all Ks from the `SC3_K` parameter)

The names of reduces dimensionality matrices (which can be retrieved by `SingleCellExperiment::reducedDim()`):

- PCA: `"pca"`
- t-SNE: `"tsne"`
- UMAP: `"umap"`

Or you can view their names with `SingleCellExperiment::reducedDimNames()`.

## Selection of PCs

`pca_elbow_pcs`, `pca_gene_var_pcs`: a number of selected PCs for elbow point and technical variance methods, respectively.

`pca_selected_pcs_plot`: a `{ggplot2}` object showing cummulative variance explained and the selected
number of first PCs for each of the methods.

## Cell clustering

`clusters_all`: a named list of integer vectors which wraps all clusterings below.

### Graph-based clustering

`graph_k`: an `{igraph}` object of shared nearest-neighbor graph, returned from `scran::buildSNNGraph()`.

`cluster_graph_walktrap`, `cluster_graph_louvain`: lists with cell clusters returned from `igraph::cluster_walktrap()`
and `igraph::cluster_louvain()`, respectively. Both functions are using `graph_k` as the input.

`cluster_graph_walktrap_n`, `cluster_graph_louvain_n`: a number of found clusters.

`cluster_graph_walktrap_table`, `cluster_graph_louvain_table`: a dataframe summarizing assignment of cells to clusters.

### k-means

`kmeans_gaps`: a `clusGap` object used to calculate the best K. Returned from `cluster::clusGap()`, to which matrix of
selected PCs is passed.

`kmeans_best_k`: the best K for k-means, returned from `cluster::maxSE()`.

`cluster_kmeans_kbest`, `cluster_kmeans_kc`: lists with cell clusters returned from `stats::kmeans()`.
In the former, `kmeans_best_k` is used as the number of clusters. In the latter, Ks from the `KMEANS_K` parameter are used.

`cluster_kmeans_kbest_table`, `cluster_kmeans_kc_tables`: a dataframe (or list of them in the latter) summarizing
assignment of cells to clusters.

### SC3

`sce_sc3`: a `SingleCellExperiment` object returned from `SC3::sc3()`.

`cluster_sc3`: a list with cell clusters.

`cluster_sc3_tables`: a list of dataframes, each summarizing assignment of cells to clusters for different `SC3_K`.

`cluster_sc3_stability_plots`: a list of `{ggplot2}` objects, each returned from `make_sc3_stability_plots()`
(where `SC3::sc3_plot_cluster_stability()` is used internally). If the parameter `SC3_DRY` is `True`, the plots
will be empty and will contain the text "SC3 was run in dry mode.".

## Plots

`hvg_plot`: a `{patchwork}` object with HVG metric statistics (average expression vs. variance).

***

`pca_phase_plots`: a list of one or two `{ggplot2`} objects. The first one is a plot of the first two PCs colored by
cell cycle phase, where PCA was computed using HVGs. If the parameter `HVG_RM_CC_GENES` is `True`,
the second plot is the same as the first one, but with PCA computed on all genes. In that case it also means that
in the first plot, PCA was computed using HVGs with removed cell cycle-related genes.

***

`pca_doublet_plot`, `pca_total_plot`: a `{ggplot2`} plots of the first two PCs colored by doublet score and total
number of UMIs, respectively.

***

`dimred_plots_clustering`: a tibble holding plots of reduced dimensions colored by a clustering.

***

`dimred_plots_other_vars`: a tibble holding plots of reduced dimensions colored by a specified variable,
as defined by the `NORM_CLUSTERING_REPORT_DIMRED_NAMES` and `NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER` parameters.

### Selected markers

`selected_markers_plots`: a tibble holding selected marker plots (a `{patchwork}` objects) for each dimemnsional
reduction specified in the `NORM_CLUSTERING_REPORT_DIMRED_NAMES` parameter. Selected markers are read from CSV file
defined in the `SELECTED_MARKERS_FILE` parameter.

You can also use the underlying function `selected_markers_dimplot()`:

```{r, eval = FALSE}
selected_markers_dimplot(
  sce = sce_final_norm_clustering,
  dimred = "umap",
  selected_markers_df = selected_markers_df,
  assay = "logcounts"
)
```

The `selected_markers_df` tibble can be read from the CSV file with

```{r, eval = FALSE}
selected_markers_df <- readr::read_csv("markers.csv", col_names = c("group", "markers"), col_types = "cc")
```

## Other targets

`config_norm_clustering`: a list holding parameters for this stage.

***

`cc_genes`: a dataframe of cell cycle genes, based on `Seurat::cc.genes.updated.2019` data. If the `ORGANISM` parameter
in the `00_main.yaml` config is `"mouse"`, then gene symbols are first converted to sentence-case (`"MKI67"` -> `"Mki67"`).

***

`cell_data`: a DataFrame holding `colData()` of `sce_rm_doublets`, plus clusterings from the `clusters_all` target,
and the new cell groupings defined in the `CELL_GROUPINGS` parameter.
