---
title: "Normalization and clustering stage (`02_norm_clustering`)"
description: "02_norm_clustering stage of the single-sample pipeline."
date: "`r glue::glue('<sup>Document generated: {format(Sys.time(), \"%Y-%m-%d %H:%M:%S %Z%z</sup>\")}')`"
package: scdrake
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Normalization and clustering stage (`02_norm_clustering`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

***

## {.tabset}

### Overview

This stage build upon the `01_input_qc` stage and performs perhaps the most important steps in scRNA-seq data analysis:
normalization, highly variable gene selection, dimensionality reduction, clustering, cell type annotation, and
visualization.

`r emoji::emoji("gear")` Config file: `config/single_sample/02_norm_clustering.yaml`

`r emoji::emoji("clipboard")` HTML report target (in `config/pipeline.yaml`):
`DRAKE_TARGETS: ["report_norm_clustering", "report_norm_clustering_simple"]`

[`r emoji::emoji("scroll")` Example report](https://onco.img.cas.cz/novotnyj/scdrake/single_sample/pbmc1k/02_norm_clustering/02_norm_clustering.html)
([used config](https://github.com/bioinfocz/scdrake/blob/main/tests/testthat/run_pipeline_config_patches/single_sample_pbmc1k_full/02_norm_clustering.default.yaml))

[`r emoji::emoji("scroll")` Example simplified report](https://onco.img.cas.cz/novotnyj/scdrake/single_sample/pbmc1k/02_norm_clustering/02_norm_clustering_simple.html)
([used config](https://github.com/bioinfocz/scdrake/blob/main/tests/testthat/run_pipeline_config_patches/single_sample_pbmc1k_full/02_norm_clustering.default.yaml))

`r emoji::emoji("ladder")` Structure

- Calculation of cell cycle score and assignment of cells to phases (G1, G2M, S) using the list of cell cycle genes in
  `Seurat::cc.genes.updated.2019` and `Seurat::CellCycleScoring()` function
- [Normalization](https://bioconductor.org/books/3.15/OSCA.basic/normalization.html) of UMI counts by either:
  - [Normalization by deconvolution](https://bioconductor.org/books/3.15/OSCA.basic/normalization.html#normalization-by-deconvolution)
    implemented in the [scran](https://bioconductor.org/packages/3.15/bioc/html/scran.html) package
    (`scuttle::computePooledFactors()`, `scuttle::logNormCounts()`)
  - Regularized negative binomial regression to normalize UMI count data implemented in the
    [sctransform](https://github.com/ChristophH/sctransform) package and wrapped by the
    [Seurat](https://satijalab.org/seurat/) package. This method also allows to regress out confounding variables
    such as cell cycle score or percentage of mitochondrial genes expression. Note that for cell cycle score,
    a better method seems to be removal of CC-related genes prior to detection of highly variable genes (HVGs, see below).
    `sctransform` also returns a specified number of HVGs.
- Selection of [highly variable genes](https://bioconductor.org/books/3.15/OSCA.basic/feature-selection.html) (HVGs):
  - HVGs are used downstream for dimensionality reduction and clustering
  - Three metrics for HVGs are available:
    - [Gene variance](https://bioconductor.org/books/3.15/OSCA.basic/feature-selection.html#quantifying-per-gene-variation)
    - [Coefficient of variation](http://bioconductor.org/books/3.15/OSCA.advanced/more-hvgs.html#using-the-coefficient-of-variation)
      (CV2)
    - HVGs obtained from the `sctransform` normalization
  - Based on the selected one from the first two metrics above, HVGs are selected either by:
    - Top N genes based on a metric (e.g. top 1000 genes with the highest variance)
    - Significance (FDR) threshold
    - Threshold on a metric value
  - Prior to HVG selection, cell cycle-related genes can be removed.
    `{scdrake}` is using a method based on the percentage of variance explained by the cell cycle phase
    in the expression profile for each gene
    ([details](https://bioconductor.org/books/3.15/OSCA.advanced/cell-cycle-assignment.html#removing-cell-cycle-related-genes))
- Removal of [cell doublets](https://bioconductor.org/books/3.15/OSCA.advanced/doublet-detection.html)
  (`scDblFinder::computeDoubletDensity()`).
- [Calculation of PCA](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#principal-components-analysis)
  and [selection of number of principal components](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#choosing-the-number-of-pcs)
  (PCs) which will be used downstream. There are three methods to select a proper number of PCs:
  - Using the [elbow point](http://bioconductor.org/books/3.15/OSCA.advanced/dimensionality-reduction-redux.html#more-choices-for-the-number-of-pcs)
  - Using the [technical noise](http://bioconductor.org/books/3.15/OSCA.advanced/dimensionality-reduction-redux.html#using-the-technical-noise)
  - Using a fixed number of PCs (`forced`)
- [Cell clustering](https://bioconductor.org/books/3.15/OSCA.basic/clustering.html) using several algorithms:
  - [Graph-based clustering](https://bioconductor.org/books/3.15/OSCA.basic/clustering.html#clustering-graph)
    (mutual nearest neighbors) using Leiden, Louvain or Walktrap algorithms.
    To control for the number of resulting clusters (granularity), multiple resolutions can be specified for the former two algorithms.
  - [K-means clustering](https://bioconductor.org/books/3.15/OSCA.basic/clustering.html#vector-quantization-with-k-means)
    using both defined numbers of K and best K selection
  - [Single-Cell Consensus Clustering](https://bioconductor.org/packages/3.15/bioc/html/SC3.html) (SC3) with
    defined numbers of clusters
- Reference-based cell type annotation using `{SingleR}`. More details are given in the accompanying
  [book](http://bioconductor.org/books/3.15/SingleRBook/) and the underlying method is described
  [here](http://bioconductor.org/books/3.15/SingleRBook/introduction.html#method-description).
  - Reference datasets are taken by default from the `{celldex}` package.
  - Note that the reference datasets included in the `{celldex}` package don't contain every possible cell type and the
  annotation may bring inaccurate labels for some datasets. Thus, it is also possible to supply `{SingleR}` a custom
  reference datasets in the form of `{SingleCellExperiment}` or `{SummarizedExperiment}` object.
  - Plots of annotation diagnostics:
    - Score heatmaps show distribution of predicted cell types in computed clusters (if any), along with per-cell annotation scores,
    - Marker heatmaps show genes that are markers for a given cell type in both the reference and current datasets,,
      i.e. those markers have driven the decision to label cells by the chosen cell type,
    - Delta scores show poor-quality or ambiguous assignments based on the per-cell 'delta', i.e., the difference between,
      the score for the assigned label and the median across all labels for each cell.,
      See [OSCA](https://bioconductor.org/books/3.15/SingleRBook/annotation-diagnostics.html#based-on-the-deltas-across-cells) for more details
- Assignment of cells to predefined groupings based on existing data. That means you can reuse an assignment of cells
  from e.g. clustering, and rename or merge its levels (cluster numbers).
  - Example: you identify a biological meaning of clusters computed by k-means with $k = 3$.
    You can use this information, and rename and merge the clusters, such that:
    - Cluster 1 -> T cells
    - Cluster 2 -> B cells
    - Cluster 3 -> T cells
  - Then you can use the new cell grouping for e.g. plotting of reduced dimensions or calculation of cluster markers.
- [Dimensionality reduction](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html):
  - [PCA](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#principal-components-analysis)
  - [t-SNE](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#t-stochastic-neighbor-embedding)
  - [UMAP](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#uniform-manifold-approximation-and-projection)
- Expression plots of selected groups of genes (markers) in reduced dimensions
- A simplified report for non-technical audience which contains only dimred plots, selected markers and cell annotation results

### Config parameters

Config for this stage is stored in the `config/single_sample/02_norm_clustering.yaml` file.
Directory with this file is read from `SCDRAKE_SINGLE_SAMPLE_CONFIG_DIR` environment variable upon `{scdrake}` load or
attach, and saved as `scdrake_single_sample_config_dir` option. This option is used as the default argument value in
several `{scdrake}` functions.

***

#### Normalization parameters

```yaml
NORMALIZATION_TYPE: "scran"
```

**Type:** character scalar (`"scran"` | `"sctransform"` | `"none"`)

A normalization type:

- `"scran"`: normalization by deconvolution.
- `"sctransform"`: regularized negative binomial regression. The returned counts in log1p (natural log) scale are
  converted to log2. Note: `SingleCellExperiment` normalized by this method cannot be used for integration.
- `"none"`: skip normalization. This should be used only in case you are importing a `SingleCellExperiment`
  object in the `01_input_qc` stage and this object was normalized before within this pipeline,
  i.e. you are importing a modified `sce_norm` target or one of its downstream, dependent `SingleCellExperiment` targets.
  - For example, you can take `sce_final_norm_clustering`, subset cells to e.g. three clusters, save the result to Rds
    file, and run the pipeline again while importing this file in the `01_input_qc` stage.
    Note that in that case you should disable empty droplets removal, and cell and gene filtering,
    as data are already of high quality. Overall, this is an example of reclustering, which is otherwise hard to
    achieve within the pipeline.

See `vignette("pipeline_overview")` for more details about normalization methods.

##### `scran` normalization parameters

These parameters are used when `NORMALIZATION_TYPE` is `"scran"`.

***

```yaml
SCRAN_USE_QUICKCLUSTER: True
```

**Type:** logical scalar

Whether to cluster cells prior to normalization. See `?scran::quickCluster` for more details.

***

```yaml
SCRAN_QUICKCLUSTER_METHOD: "igraph"
```

**Type:** character scalar (`"igraph"` | `"hclust"`)

A clustering method: `"igraph"` uses graph-based clustering, while `"hclust"` uses hierarchical clustering.
See `?scran::quickCluster` for more details.

##### `sctransform` normalization parameters

These parameters are used when `NORMALIZATION_TYPE` is `"sctransform"`.

***

```yaml
SCT_VARS_TO_REGRESS: null
```

**Type:** list of character scalar or `null`

Batch variables to regress out during the `SCTransform` normalization. You can regress out variables such as
`["percent.mt", "nFeature_RNA", "nCount_RNA"]` or cell cycle using `["S.Score", "G2M.Score"]` or `["CC.Difference"]`.

However, for correction of cell cycle effect, a better method is the removal of cell cycle-related genes during
selection of highly variable genes (HVGs).
This can be set by [HVG selection parameters](#highly-variable-genes-hvgs-selection), and the approach is explained
in more details in `vignette("pipeline_overview")`.

***

```yaml
SCT_N_HVG: 3000
```

**Type:** integer scalar

A number of top HVGs to return from the `SCTransform` normalization.

#### Highly variable genes (HVGs) selection

```yaml
HVG_METRIC: "gene_var"
```

**Type:** character scalar (`"gene_var"` | `"gene_cv2"` | `"sctransform"`)

A metric used to find HVGs. See <https://bioconductor.org/books/3.15/OSCA.basic/feature-selection.html> for more details.

- `"gene_var"`: variance of the log-counts (`scran::modelGeneVar()`).
- `"gene_cv2"`: coefficient of variation (`scran::modelGeneCV2()`).
- `"sctransform"`: use HVGs selected by `SCTransform`.
  Only relevant when the `NORMALIZATION_TYPE` parameter is `"sctransform"`.
  You can adjust number of these HVGs by the `SCT_N_HVG` parameter.

***

```yaml
HVG_SELECTION: "top"
```

**Type:** character scalar (`"top"` | `"significance"` | `"threshold"`)

A HVG selection strategy. Only relevant when the `HVG_METRIC` parameter is `"gene_var"` or `"gene_cv2"`.

- `"top"`: take top X genes according to the metric
  ([details](https://bioconductor.org/books/3.15/OSCA.basic/feature-selection.html#hvg-selection)).
- `"significance"`: use a FDR threshold
  ([details](https://bioconductor.org/books/3.15/OSCA.advanced/more-hvgs.html#more-hvg-selection-strategies)).
- `"threshold"`: use a threshold on minimum value of the metric
  ([details](https://bioconductor.org/books/3.15/OSCA.advanced/more-hvgs.html#more-hvg-selection-strategies)).

For `"top"` and `"threshold"`, `bio` and `ratio` columns are used for `HVG_METRIC` of `"gene_var"` and `"gene_cv2"`,
respectively. These columns are present in the DataFrame returned from the underlying method
(`scran::modelGeneVar()` or `scran::modelGeneCV2()`).

***

```yaml
HVG_SELECTION_VALUE: 1000
```

**Type:** depends on value of the `HVG_SELECTION` parameter:

- A positive integer scalar for `HVG_SELECTION: "top"`.
- A numeric scalar $<0; 1>$ for `HVG_SELECTION: "significance"`.
- A numeric scalar for `HVG_SELECTION: "threshold"`.

A threshold value for HVG selection. The following parameter combinations are recommended defaults:

- Top 1000 genes by a HVG metric.

```yaml
HVG_SELECTION: "top"
HVG_SELECTION_VALUE: 1000
```

- Genes with a HVG metric significance (FDR) lesser than `0.05`.

```yaml
HVG_SELECTION: "significance"
HVG_SELECTION_VALUE: 0.05
```

- Genes with a HVG metric value greater than `0` or `1`.

```yaml
HVG_METRIC: "gene_var"
HVG_SELECTION: "threshold"
HVG_SELECTION_VALUE: 0
```

```yaml
HVG_METRIC: "gene_cv2"
HVG_SELECTION: "threshold"
HVG_SELECTION_VALUE: 1
```

#### Removal of cell cycle-related genes prior to HVG selection

```yaml
HVG_RM_CC_GENES: False
```

**Type:** logical scalar

Whether to apply a correction for cell cycle prior to HVG selection.
We implement a cell cycle-related genes removal strategy described
[here](https://bioconductor.org/books/3.15/OSCA.advanced/cell-cycle-assignment.html#removing-cell-cycle-related-genes).
It is based on the percentage of variance explained by the cell cycle phase in the expression profile for each gene.

Note that when `SCTransform` is used for normalization including cell cycle variables to regress out, than
the results of this strategy remain unclear.

***

```yaml
HVG_CC_GENES_VAR_EXPL_THRESHOLD: 5
```

**Type:** positive numeric scalar

A threshold on gene cell cycle variance explained, i.e. genes with var. expl. greater than this threshold will be
marked as cell cycle-related and removed *prior* to selection of HVGs.

#### Doublet detection and filtering

Doublets are droplets in which multiple cells were captured.

***

```yaml
MAX_DOUBLET_SCORE: 3.5
```

**Type:** positive numeric scalar or `null`

A threshold for doublet score computed by `scDblFinder::computeDoubletDensity()`.
Cells with doublet score greater than the threshold will be removed.
If `null`, doublets are not removed, but doublet score is still computed.

#### Dimensionality reduction

##### PCA

```yaml
PCA_SELECTION_METHOD: "forced"
```

**Type:** character scalar (`"forced"` | `"elbow"` | `"technical_noise"`)

A method to select first N principal components (PCs):

- `"forced"`: use the number of PCs specified by the `PCA_FORCED_PCS` parameter.
- `"elbow"`: use the
  [elbow point method](https://bioconductor.org/books/3.15/OSCA.advanced/dimensionality-reduction-redux.html#using-the-elbow-point).
- `"technical_noise"`: use the
  [technical noise](https://bioconductor.org/books/3.15/OSCA.advanced/dimensionality-reduction-redux.html#using-the-technical-noise).

***

```yaml
PCA_FORCED_PCS: 15
```

**Type:** positive integer scalar

If the `PCA_SELECTION_METHOD` parameter is `"forced"`, use this number of first principal components.

##### t-SNE

[t-stochastic neighbor embedding](https://bioconductor.org/books/3.15/OSCA.basic/dimensionality-reduction.html#t-stochastic-neighbor-embedding)

See `?scater::runTSNE` for the underlying method.

***

```yaml
TSNE_PERP: 20
```

**Type:** numeric scalar

t-SNE perplexity.

***

```yaml
TSNE_MAX_ITER: 1000
```

**Type:** positive integer scalar

A maximum number of t-SNE iterations.

#### Clustering

##### Graph-based clustering

Parameters for [graph-based clustering](https://bioconductor.org/books/3.15/OSCA.basic/clustering.html#clustering-graph).

***

```yaml
CLUSTER_GRAPH_SNN_K: 10
```

**Type:** positive integer scalar

A number of shared nearest neighbors in graph used for community detection algorithms (graph-based clustering).
Passed as the `k` parameter to `bluster::makeSNNGraph()`. From the function's help page:

> The choice of k controls the connectivity of the graph and the resolution of community detection algorithms.
Smaller values of k will generally yield smaller, finer clusters, while increasing k will increase the connectivity of
the graph and make it more difficult to resolve different communities.
The value of k can be roughly interpreted as the anticipated size of the smallest subpopulation.

***

```yaml
CLUSTER_GRAPH_SNN_TYPE: "rank"
```

**Type:** character scalar (`"rank"` | `"number"` | `"jaccard"`)

A type of weighting scheme to use for shared neighbors. Passed as `type` parameter to `bluster::makeSNNGraph()`
(see the function's help page for more information).

***

```yaml
CLUSTER_GRAPH_LEIDEN_ENABLED: True
```

**Type:** logical scalar

If `True`, enable Leiden community detection algorithm.

***

```yaml
CLUSTER_GRAPH_LEIDEN_RESOLUTIONS: [0.2, 0.4, 0.6, 0.8, 1.0]
```

**Type:** numeric vector

Resolutions for Leiden clustering. Lower values result in more coarse-grained clustering, while higher ones to more
fine-grained structures. Cell-cluster memberships will be named as `cluster_graph_leiden_r{r}` (e.g. in `colData()` of a SCE object).

***

```yaml
CLUSTER_GRAPH_LOUVAIN_ENABLED: False
```

**Type:** logical scalar

If `True`, enable Louvain community detection algorithm.

***

```yaml
CLUSTER_GRAPH_LOUVAIN_RESOLUTIONS: [0.2, 0.4, 0.6, 0.8, 1.0]
```

**Type:** numeric vector

Resolutions for Louvain clustering. Lower values result in more coarse-grained clustering, while higher ones to more
fine-grained structures. Cell-cluster memberships will be named as `cluster_graph_louvain_r{r}` (e.g. in `colData()` of a SCE object).

***

```yaml
CLUSTER_GRAPH_WALKTRAP_ENABLED: False
```

**Type:** logical scalar

If `True`, enable Walktrap community detection algorithm.

##### k-means clustering

```yaml
CLUSTER_KMEANS_K_ENABLED: False
```

**Type:** logical scalar

If `True`, enable k-means clustering with selected numbers of clusters (`k`s).

***

```yaml
CLUSTER_KMEANS_K: !code 3:6
```

**Type:** integer vector

A vector of number of clusters for which to compute k-means.

Note: `!code 3:6` will evaluate to `c(3, 4, 5, 6)`.

***

```yaml
CLUSTER_KMEANS_KBEST_ENABLED: False
```

**Type:** logical scalar

If `True`, enable k-means clustering with best `k` based on gap statistics.

##### SC3 clustering

```yaml
CLUSTER_SC3_ENABLED: False
```

**Type:** logical scalar

If `True`, enable SC3 with selected numbers of clusters (`k`s).

***

```yaml
CLUSTER_SC3_K: !code 5:6
```

**Type:** integer vector

A vector of number of clusters for which to compute SC3 clustering.

Note: `!code 3:6` will evaluate to `c(3, 4, 5, 6)`.

***

```yaml
CLUSTER_SC3_N_CORES: 1
```

**Type:** integer scalar

A number of CPU cores to use for SC3 computation.
This may lead to out-of-memory errors for a number of cores greater than 1.

Note that `drake` may have problems when the pipeline is run in parallel mode and at the same time a target is using
a parallel code (*within target parallelism*). For the SC3 computation is recommended to run the pipeline in
sequential mode (`DRAKE_PARALLELISM: "loop"`) and set `DRAKE_TARGETS: ["cluster_sce_sc3"]`.
Once the `cluster_sce_sc3` target is finished, you can switch the parallelism back to `clustermq` or `future`.

***

```yaml
CLUSTER_SC3_ENABLED: False
```

**Type:** logical scalar

If `False`, skip SC3 clustering.

***

#### Cell type annotation

```yaml
CELL_ANNOTATION_SOURCES_DEFAULTS:
  TRAIN_PARAMS:
    GENES: "de"
    SD_THRESH: 1
    DE_METHOD: "classic"
    DE_N: null
    ASSAY_TYPE: "logcounts"
  CLASSIFY_PARAMS:
    QUANTILE: 0.8
    TUNE_THRESH: 0.05
    ASSAY_TYPE: "logcounts"
  PRUNE_SCORE_PARAMS:
    N_MADS: 3
    MIN_DIFF_MED: -.inf
    MIN_DIFF_NEXT: 0
  DIAGNOSTICS_PARAMS:
    HEATMAP_N_TOP_MARKERS: 20
```

**Type:** named list (dictionary) of named lists

Default parameters for computation and reporting of cell type annotation. These can be overriden for each of the cell
annotation reference source in `CELL_ANNOTATION_SOURCES` (see below).

- `TRAIN_PARAMS`: used for training, passed to `SingleR::trainSingleR()`.
  - `GENES: "de"` (character scalar: `"de"` | `"sd"` | `"all"`):
    feature selection mode, see Details in `?SingleR::trainSingleR()`.
  - `SD_THRESH: 1` (numeric scalar): the minimum threshold on the standard deviation per gene.
    Only used when `GENES: "sd"`.
  - `DE_METHOD: "classic"` (character scalar: `"classic"` | `"wilcox"` | `"t"`): how DE genes should be detected
    between pairs of labels. Defaults to `"classic"`, which sorts genes by the log-fold changes and takes the top `DE_N`.
    Setting to `"wilcox"` or `"t"` will use Wilcoxon ranked sum test or Welch t-test between labels, respectively,
    and take the top `DE_N` upregulated genes per comparison.
  - `DE_N: null` (`null` or integer scalar): the number of DE genes to use when `GENES: "de"`. If `DE_METHOD: "classic"`,
    defaults to $500 * (2/3) ^ log2(N)$ where $N$ is the number of unique labels. Otherwise, `null` defaults to `10`.
  - `ASSAY_TYPE: "logcounts"` (character scalar or integer): the assay of the reference dataset containing the relevant
    expression matrix.
- `CLASSIFY_PARAMS`: used for classification, passed to `SingleR::classifySingleR()`.
  - `QUANTILE: 0.8` (numeric scalar): the quantile of the correlation distribution to use to compute the score for each label.
  - `TUNE_THRESH: 0.05` (numeric scalar): the maximum difference from the maximum correlation to use in fine-tuning.
  - `ASSAY_TYPE: "logcounts"` (character scalar or integer): the assay of the test dataset containing the relevant
    expression matrix.
- `PRUNE_SCORE_PARAMS`: used for label pruning after classification, passed to `SingleR::pruneScores()`.
  - `N_MADS: 3` (numeric scalar): the number of MADs to use for defining low outliers in the per-label distribution
    of delta values (i.e., difference from median).
  - `MIN_DIFF_MED: -.inf` (numeric scalar): the minimum acceptable delta for each cell.
  - `MIN_DIFF_NEXT: 0` (numeric scalar): the minimum difference between the best score and the next best score in fine-tuning.
- `DIAGNOSTICS_PARAMS`: parameters for post-classification diagnostics.
  - `HEATMAP_N_TOP_MARKERS: 20` (integer scalar): how many top markers to put into heatmaps after classification.
    Markers are computed after classification using the predicted cell labels.

***

```yaml
CELL_ANNOTATION_SOURCES:
  - human_primary_cell_atlas_main:
      reference_type: "celldex"
      reference: "HumanPrimaryCellAtlasData"
      description: >
        Microarray datasets derived from human primary cells (Mabbott et al. 2013).
        Most of the labels refer to blood subpopulations but cell types from other tissues are also available.
      label_column: "label.main"
      label_subsets: []
    monaco_immune_fine:
      reference_type: "celldex"
      reference: "MonacoImmuneData"
      description: "This is the human immune reference that best covers all of the bases for a typical PBMC sample."
      label_column: "label.fine"
      label_subsets: []
```

**Type:** list of named lists

Specifies reference dataset(s) to be used for cell type annotation. The method used in `{SingleR}` for training and
classification is described [here](http://bioconductor.org/books/3.15/SingleRBook/introduction.html#method-description).

The names of named lists are used to name the reference datasets, e.g. as the prefixes in names of `colData()` columns
containing per-cell labels.
More info about the structure of related targets is given in the [Targets](#cell-type-annotation-1) section.

Each reference dataset must have the following parameters:

- `reference_type: "celldex"` (character scalar: `"celldex"` | `"file"`): type of the reference dataset.
- `reference: "HumanPrimaryCellAtlasData"` (character scalar):
  - If `reference_type: "celldex"`: a name of function to call in the `{celldex}` package. This function must return a
    `SummarizedExperiment` object. See `{celldex}`'s
    [vignette](https://bioconductor.org/packages/3.15/data/experiment/vignettes/celldex/inst/doc/userguide.html)
    for an overview of available datasets.
  - If `reference_type: "file"`: a path to Rds file with saved `SingleCellExperiment` or `SummarizedExperiment` object.
    The assay type specified `TRAIN_PARAMS/assay_type` must be present in `assayNames()` of the object.
    The `label_column` must be present in `colData()` of the object.
- `description: "..."` (character scalar): a description of the reference dataset, will appear in some plots.
- `label_column: "label.main"` (character scalar): specifies which column from `colData()` of the reference dataset
  will be used for training. For datasets from the `{celldex}` package, those are `label.main`,
  `label.fine` (more granular labels), or `label.ont` (cell ontology IDs).
- `label_subsets: []` (character vector): labels to subset the reference dataset prior to training,
  e.g. `["T cells", "B cells", "Progenitors"]`.

In each entry (e.g. `human_primary_cell_atlas_main` above) you can override default parameters in
`CELL_ANNOTATION_SOURCES_DEFAULTS` (by using lowercase names), e.g.

```yaml
CELL_ANNOTATION_SOURCES:
  - human_primary_cell_atlas_main:
      ...
      train_params:
        de_method: "wilcox"
        de_n: 30
```

will override corresponding parameters in `TRAIN_PARAMS` in `CELL_ANNOTATION_SOURCES_DEFAULTS`.

***

#### Cell grouping assignment

Using the two parameters below, it is possible to add custom cell metadata or rearrange the current one into new groupings
(saved as `colData()` columns of the `SingleCellExperiment` object). These groupings can be then referenced in other parameters,
e.g. `NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER` or in the `cluster_markers` and `contrasts` stages.

```yaml
ADDITIONAL_CELL_DATA_FILE: null
```

**Type:** character scalar or `null`

A path to file with additional cell data which will be added to `colData()` of the final SCE object of this stage.
The additional data will overwrite existing columns (e.g. `cluster_graph_louvain`), but are appended before the
`CELL_GROUPINGS` below is applied, so you can refer to additional data columns inside `CELL_GROUPINGS`.
In the additional data, there can be only cell barcodes that are present after cell filtering in the `01_input_qc` stage -
the missing ones will have `NA` set in the resulting data. Rownames are not mandatory.

The supported file formats are CSV (comma separated, `.csv` extension) and Rds (dataframe-like objects, `.Rds` extension).
The additional data must contain `Barcode` column by which they will be joined with the original cell data (`colData()`).

***

```yaml
CELL_GROUPINGS: null
```

**Type:** list of named lists or `null`

This parameter specifies new cell groupings based on existing columns (primarily the clustering ones). Each key is a
name of new cell grouping (column in `colData()` of the `sce_dimred` target) and contains a named list in the form:

- `source_column` (character scalar): a name of existing column in `colData()`.
- `description` (character scalar, optional): a description of the cell grouping, will appear in plots.
  If not set, the name of the assignment will be used.
- `assignments` (named list/vector of character scalars):
  - `<old level>: "<new level>"`: assignment of an old level to a new level. Different old levels with the same new
    level will be merged. Order of assignments does not matter, levels are just replaced
    (under the hood, `dplyr::recode()` is used). Unspecified levels will be kept as they are.
    You cannot replace existing groupings (columns).

For possible column names for `source_column`, see the `sce_final_norm_clustering` target in the [section](#targets) below.

Let's look at example cell grouping:

```yaml
CELL_GROUPINGS:
  - cluster_graph_leiden_r0.4_annotated:
      source_column: "cluster_graph_leiden_r0.4"
      description: "Graph-based clustering (Leiden alg.), annotated clusters"
      assignments:
        1: "memory_CD4+"
        6: "B"
        7: "memory_CD4+"
```

- `cluster_graph_leiden_r0.4_annotated`: a name of new cell grouping added to `colData()`.
  - `source_column: "cluster_graph_leiden_r0.4"`: a name of column in `colData()` whose level will be used for the
    new assignments.
  - `description: "Graph-based clustering ..."`: a description of the new assignment. If it was not set,
    the name of the grouping (`cluster_graph_leiden_r0.4_annotated`) would be used instead.
  - `assignments`: assignments of the old levels to new ones. That is, level (cluster) `6` will be renamed to `"B"`,
    and levels `1` and `7` will be merged to a new level `"memory_CD4+"`.
    All unspecified levels will be kept as they are.

#### Dimensionality reduction plots

```yaml
NORM_CLUSTERING_REPORT_DIMRED_NAMES: ["umap", "pca", "tsne"]
```

**Type:** character vector (`"umap"` | `"pca"` | `"tsne"`)

A vector of names of dimensionality reduction methods for which plots in the report will be made.

***

```yaml
NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER:
  - "phase": "Cell cycle phases"
    "doublet_score": "Doublet score"
    "total": "Total number of UMI"
    "detected": "Detected number of genes"
```

**Type:** list of named lists with character scalars

Names of other variables in `colData()` to plot dimreds and color by, displayed in the report.

The format is `"variable_name": "description"`, where description will appear in plot title.

Variables defined in the `CELL_GROUPINGS` parameter can have `description` set to `null` -
their description from `CELL_GROUPINGS` will be used.

#### Input files

```yaml
SELECTED_MARKERS_FILE: null
```

**Type:** character scalar or `null`

A path to CSV file with groups of markers to make expression dimred plots for. Set to `null` to skip.
The plots will be created for each of `NORM_CLUSTERING_REPORT_DIMRED_NAMES`.

The format is following: each row defines gene symbols for a group.
First column is a name of group and second column are gene symbols delimited by ":". Example:

`Naive_CD4+_T,IL7R:CCR7`

**Do not use a header!**

An example CSV file for PBMC 1k dataset is automatically copied to a newly initialized project
(as `system.file("extdata", "selected_markers.csv", package = "scdrake", mustWork = TRUE)`).
You can also view this file on GitHub [here](https://github.com/bioinfocz/scdrake/blob/main/inst/extdata/selected_markers.csv).

***

```yaml
NORM_CLUSTERING_REPORT_RMD_FILE: "Rmd/single_sample/02_norm_clustering.Rmd"
NORM_CLUSTERING_REPORT_SIMPLE_RMD_FILE: "Rmd/single_sample/02_norm_clustering_simple.Rmd"
```

**Type:** character scalar

Paths to RMarkdown files used for HTML reports of this pipeline stage.

`NORM_CLUSTERING_REPORT_RMD_FILE` is technically more detailed, while `NORM_CLUSTERING_REPORT_SIMPLE_RMD_FILE`
outputs only the dimensionality reduction plots along with cell annotation.

#### Output files

```yaml
NORM_CLUSTERING_BASE_OUT_DIR: "02_norm_clustering"
```

**Type:** character scalar

A path to base output directory for this stage. It will be created under `BASE_OUT_DIR` specified in `00_main.yaml` config.

***

```yaml
NORM_CLUSTERING_SELECTED_MARKERS_OUT_DIR: "selected_markers"
NORM_CLUSTERING_DIMRED_PLOTS_OUT_DIR: "dimred_plots"
NORM_CLUSTERING_OTHER_PLOTS_OUT_DIR: "other_plots"
NORM_CLUSTERING_CELL_ANNOTATION_OUT_DIR: "cell_annotation"
NORM_CLUSTERING_REPORT_HTML_FILE: "02_norm_clustering.html"
NORM_CLUSTERING_REPORT_SIMPLE_HTML_FILE: "02_norm_clustering_simple.html"
```

**Type:** character scalar

Names of files and directories created under `NORM_CLUSTERING_BASE_OUT_DIR`. Subdirectories are not allowed.

#### HTML output parameters

```yaml
NORM_CLUSTERING_KNITR_MESSAGE: False
NORM_CLUSTERING_KNITR_WARNING: False
NORM_CLUSTERING_KNITR_ECHO: False
```

**Type:** logical scalar

These are passed to `knitr::opts_chunk()` and used for rendering of stage's HTML report.

### Outputs

Here you can find description of the most important targets for this stage.
However, for a full overview, you have to inspect the
[source code](https://github.com/bioinfocz/scdrake/blob/main/R/plans_single_sample.R) of the
`get_input_norm_clustering_subplan()` function.

HTML report target names: `report_norm_clustering`, `report_norm_clustering_simple` (only dimred plots)

#### `SingleCellExperiment` objects

`sce_cc`: a SCE object (the `sce_final_input_qc` target from the `input_qc` stage) with added cell cycle information
to `colData()`: `phase` (`"G1"`, `"G2m"` or `"S"`), `s_score`, `g2m_score`, `cc_difference` (`s_score` - `g2m_score`).

***

`sce_norm`: `sce_cc` with normalized counts in `SingleCellExperiment::logcounts()` assay and `normalization_type` item
added to `metadata()`.

***

`sce_norm_hvg`: `sce_norm` with added HVG information:

- `metadata()`: `hvg_metric`, `hvg_selection`, `hvg_selection_value`, `hvg_metric_fit`, `hvg_rm_cc_genes`, `hvg_ids`
- `rowData()`: `is_hvg` (logical)

If the parameter `HVG_RM_CC_GENES` is `True`, `sce_norm_hvg` will also contain a PCA dimensionality reduction
(named `pca_with_cc`) for whose computation all HVGs were used.

***

`sce_rm_doublets`: `sce_norm_hvg` with removed doublets and added information:

- `metadata()`: `has_filtered_doublets` (`FALSE` if `MAX_DOUBLET_SCORE` parameter is `null`), `max_doublet_score`
- `colData()`: `doublet_score` (numeric), `is_doublet` (logical)

***

`sce_pca`: `sce_rm_doublets` with calculated PCA (50 PCs). Reduced dimensions can be retrieved by
`SingleCellExperiment::reducedDim(sce_pca, "pca")`

***

`sce_pca_selected_pcs`: `sce_pca` with selected number of PCs. Reduced dimensions in `pca` slot are subsetted to
the selected number of PCs, and `pca_all` slot contains the full matrix of 50 PCs. Information about selection of PCs
is added to `metadata()`: `pca_selection_method`, `pca_selected_pcs`.

***

`sce_dimred`: `sce_pca_selected_pcs` with calculated t-SNE and UMAP,
both using the selected number of PCs for calculation

***

`sce_final_norm_clustering`: a final SCE object of this stage which will be used in the `cluster_markers` and
`contrasts` stages. This object is derived from `sce_dimred`, but cell clusterings along with `CELL_GROUPINGS` are
added to `colData()`.

The names of columns containing cell clusterings are:

- Graph-based clustering: `cluster_graph_walktrap`, `cluster_graph_louvain_r<r>` (e.g. `cluster_graph_louvain_r0.4`),
  `cluster_graph_leiden_r<r>` (e.g. `cluster_graph_leiden_r0.4`)
  - `<r>` is based on the `CLUSTER_GRAPH_LEIDEN_RESOLUTIONS` and `CLUSTER_GRAPH_LOUVAIN_RESOLUTIONS` parameters
- k-means clustering: `cluster_kmeans_kbest`, `cluster_kmeans_k<k>` (e.g. `cluster_kmeans_k5`)
  - `<k>` is based on the `CLUSTER_KMEANS_K` parameter
- SC3 clustering: `cluster_sc3_<k>` (e.g. `cluster_sc3_k5`)
  - `<k>` is based on the `CLUSTER_SC3_K` parameter

The names of reduces dimensionality matrices (which can be retrieved by `SingleCellExperiment::reducedDim()`):

- PCA: `"pca"`
- t-SNE: `"tsne"`
- UMAP: `"umap"`

Or you can view their names with `SingleCellExperiment::reducedDimNames()`

#### Selection of PCs

`pca_elbow_pcs`, `pca_gene_var_pcs`: a number of selected PCs for elbow point and technical variance methods, respectively

`pca_selected_pcs_plot`: a `{ggplot2}` object showing cummulative variance explained and the selected
number of first PCs for each of the methods

#### Cell clustering

`clusters_all`: a named list of integer vectors which wraps all clusterings below

##### Graph-based clustering

`graph_snn`: an `{igraph}` object of shared nearest-neighbor graph, returned from `scran::buildSNNGraph()`

`cluster_graph_leiden`, `cluster_graph_walktrap`, `cluster_graph_louvain`: lists with cell clusters returned from `igraph::cluster_walktrap()`
and `igraph::cluster_louvain()`, respectively. Both functions are using `graph_snn` as the input

`cluster_graph_louvain_df`, `cluster_graph_walktrap_df`, `cluster_graph_leiden_df`: dataframes with graph-based clustering results

`cluster_graph_leiden_clustree`, `cluster_graph_louvain_clustree`: `{ggplot2}` objects of `clustree::clustree()` plots

##### k-means

`cluster_kmeans_kbest_gaps`: a `clusGap` object used to calculate the best K. Returned from `cluster::clusGap()`, to which matrix of
selected PCs is passed

`cluster_kmeans_kbest_k`: the best K for k-means, returned from `cluster::maxSE()`

`cluster_kmeans_kbest`, `cluster_kmeans_k`: lists with cell clusters returned from `stats::kmeans()`.
In the former, `cluster_kmeans_kbest_k` is used as the number of clusters. In the latter, Ks from the `CLUSTER_KMEANS_K` parameter are used.

`cluster_kmeans_k_df`, `cluster_kmeans_kbest_df`: dataframes with k-means clustering results

`cluster_kmeans_k_clustree`: `{ggplot2}` object of `clustree::clustree()` plot. If `CLUSTER_KMEANS_KBEST_ENABLED` is `True`,
it will also contain k-means result with the best `k`.

##### SC3

`sce_sc3`: a `SingleCellExperiment` object returned from `SC3::sc3()`

`cluster_sc3`: a list with cell clusters

`cluster_sc3_df`: a dataframe with SC3 clustering results

`cluster_sc3_stability_plots`: a list of `{ggplot2}` objects, each returned from `make_sc3_stability_plots()`
(where `SC3::sc3_plot_cluster_stability()` is used internally)

`cluster_sc3_clustree`: `{ggplot2}` object of `clustree::clustree()` plot

***

#### Cell type annotation

`cell_annotation_params`: a tibble with parameters parsed from `CELL_ANNOTATION_SOURCES`

`cell_annotation`: `cell_annotation_params` with added `cell_annotation` list column holding outputs (`DataFrame`s)
from `SingleR::SingleR()`. The structure of those `DataFrame`s is described in `?SingleR::classifySingleR()`.

***

`cell_annotation_labels`: a named list of character vectors with per-cell label assignments.
Later added to the `cell_data` target and in turn, to `colData()` of the `sce_final_norm_clustering` and
`sce_int_clustering_final` (in integration pipeline) targets/objects.

By default, there are three vectors with labels for each reference dataset named as:

- `<reference_name>_labels_raw`: "raw" labels. Taken from `first.labels` column of `SingleR::SingleR()` output
- `<reference_name>_labels`: fine-tuned labels. Taken from `labels` column of `SingleR::SingleR()` output
- `<reference_name>_labels_pruned`: fine-tuned and pruned labels (contains `NA`s for low quality labels).
  Taken from `pruned.labels` column of `SingleR::SingleR()` output.

Note that you can use names of `cell_annotation_labels`'s items in the `CELL_GROUPINGS` and
`NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER` parameters, and also use them in `cluster_markers`
(`CLUSTER_MARKERS_SOURCES`) and `contrasts` (`CONTRASTS_SOURCES`) stages

***

`cell_annotation_diagnostic_plots`: `cell_annotation` with added list columns holding diagnostic plots,
i.e. for each reference dataset:

- `score_heatmaps`: heatmaps of per-cell label scores created for each clustering (as column annotation)
  ([details](http://bioconductor.org/books/3.15/SingleRBook/annotation-diagnostics.html#based-on-the-scores-within-cells)).
- `marker_heatmaps`: `NULL` if `cell_annotation$train_params$de` is not `"de"`, otherwise heatmap for each label
  containing top upregulated markers from pairwise t-tests
  ([details](http://bioconductor.org/books/3.15/SingleRBook/annotation-diagnostics.html#based-on-marker-gene-expression)).
  Number of top markers is specified in `cell_annotation$diagnostics_params$heatmap_n_top_markers`.
- `delta_distribution_plot`: violin plots (in one figure / object) of per-cell deltas for each label.
  Deltas are differences between the score for the assigned label and the median across all labels for each cell
  ([details](http://bioconductor.org/books/3.15/SingleRBook/annotation-diagnostics.html#based-on-the-deltas-across-cells)).

`cell_annotation_diagnostic_plots_files`: paths to PDF files of diagnostic plots in `cell_annotation_diagnostic_plots`
(`score_heatmaps_out_file`, `delta_distribution_plot_out_file`, `marker_heatmaps_out_file`).
That means when you make this target, those files will be created.

#### Plots

`hvg_plot`: a `{patchwork}` object with HVG metric statistics (average expression vs. variance)

***

`pca_phase_plots`: a list of one or two `{ggplot2`} objects. The first one is a plot of the first two PCs colored by
cell cycle phase, where PCA was computed using HVGs. If the parameter `HVG_RM_CC_GENES` is `True`,
the second plot is the same as the first one, but with PCA computed on all genes. In that case it also means that
in the first plot, PCA was computed using HVGs with removed cell cycle-related genes.

***

`pca_doublet_plot`, `pca_total_plot`: a `{ggplot2`} plots of the first two PCs colored by doublet score and total
number of UMIs, respectively

***

`dimred_plots_clustering`: a tibble holding plots of reduced dimensions colored by a clustering

`dimred_plots_clustering_files_out`, `dimred_plots_clustering_united_files`: make this target to export the plots
(the latter produces a single multipage PDF)

***

`dimred_plots_other_vars`: a tibble holding plots of reduced dimensions colored by a specified variable,
as defined by the `NORM_CLUSTERING_REPORT_DIMRED_NAMES` and `NORM_CLUSTERING_REPORT_DIMRED_PLOTS_OTHER` parameters.

`dimred_plots_other_vars_files_out`: make this target to export the plots

##### Selected markers

`selected_markers_plots`: a tibble holding selected marker plots (`{patchwork}` objects) for each dimensional
reduction specified in the `NORM_CLUSTERING_REPORT_DIMRED_NAMES` parameter. Selected markers are read from CSV file
defined in the `SELECTED_MARKERS_FILE` parameter.

`selected_markers_plots_files_out`: make this target to export the plots

You can also use the underlying function `selected_markers_dimplot()`:

```{r, eval = FALSE}
selected_markers_dimplot(
  sce = sce_final_norm_clustering,
  dimred = "umap",
  selected_markers_df = selected_markers_df,
  assay = "logcounts"
)
```

The `selected_markers_df` tibble can be read from the CSV file with

```{r, eval = FALSE}
selected_markers_df <- readr::read_csv("markers.csv", col_names = c("group", "markers"), col_types = "cc")
```

#### Other targets

`config_norm_clustering`: a list holding parameters for this stage

***

`cc_genes`: a dataframe of cell cycle genes, based on `Seurat::cc.genes.updated.2019` data. If the `ORGANISM` parameter
in the `00_main.yaml` config is `"mouse"`, then gene symbols are first converted to sentence-case (`"MKI67"` -> `"Mki67"`).

***

`cell_data`: a DataFrame holding `colData()` of `sce_rm_doublets`, clusterings from the `clusters_all` target,
cell annotation labels from the `cell_annotation_labels` target, and the new cell groupings defined in
the `CELL_GROUPINGS` parameter.
